---
title: "Estimating spawning salmon abundance from visual surveys: a hierarchical model of arrival and exit dynamics"
author:
  - name: Scott A. Akenhead 
    email: scott@s4s.com 
    affiliations: 
        - id: pbs 
          name: Pacific Biological Station
          department: Fisheries and Oceans Canada
          address: 3190 Hammond Bay Road
          city: Nanaimo
          state: BC, Canada
          postal-code: V9T 6N7
    note: (retired).
  - name: Colette Louie
    email: CLouie@oib.ca
    affiliations:
        - id: OIB
          name: Osooyos Indian Band
          address: 1155 Sen Pok Chin Blvd
          city: Oliver
          state: BC, Canada
          postal-code: V0H 1T3
  - name: Patrick L. Thompson
    email: patrick.thompson@dfo-mpo.gc.ca
    affiliations:
        - id: psec
          name: Pacific Science Enterprise Center
          department: Fisheries and Oceans Canada
          address: 4160 Marine Dr
          city: West Vancouver
          state: BC, Canada
          postal-code: V7V 1H2
    attributes:
        corresponding: true
date: last-modified
bibliography: bibliography.bib
abstract: |
  Spawner abundance is a fundamental variable for salmon management models, but conventional methods (trapezoidal area under the curve, TAUC) do not estimate the precision of abundance and risk bias from an assumed residence time (aka survey life, to correct for observing a salmon in repeated surveys). Here we present a model that predicts visual counts of spawners by modelling the process and timing of spawner arrivals and exits from the spawning ground. Our model analyses all surveys simultaneously using a hierarchical framework, providing annual estimates of spawner abundance, arrival timing, arrival spread, exit timing, and exit spread with associated uncertainties. By directly estimating residence time and integrating uncertainty into abundance estimates, our approach improves the accuracy, transparency, and reliability of spawner abundance estimates. First we demonstrate how the model accurately recovers parameters from simulated data, then we demonstrate its application by estimating 25 years of spawner abundance of wild sockeye salmon (_Oncorhynchus nerka_) in the Okanagan River (BC, Canada). Our median estimated residence time of 10.4 days is similar to the constant 11 days previously assumed for this population, but varies by year. While our estimates of spawner abundance are similar to previous estimates that used TAUC with a constant residence time, there are deviations in some years due to year-to-year changes in residence time.

keywords: 
  - salmon
  - Oncorhynchus
  - spawner
  - spawner survey
  - Osoyoos Lake
  - Okanagan River
  - Columbia River 

format:
  elsevier-pdf:
    keep-tex: false
    number-sections: true
    journal:
      name: Fisheries Research
      formatting: preprint
      # model: 3p # Don't set a model with preprint
      cite-style: authoryear
appendix: true
execute: 
  cache: true
---

```{r setup, include = FALSE}
library(knitr)
library(magrittr)
library(tidyverse)
library(tidybayes)
library(rstan)
library(loo)
library(patchwork)
library(ggrepel)
library(janitor)
library(scales)
library(geomtextpath)
library(lubridate)
library(ggallin)
library(kableExtra)
library(ggokabeito)

theme_set(theme_bw())
```

```{r load_data, include=FALSE}
getwd()
file.exists("../data/spawn_timing.csv")

spawners <- read.csv("../data/spawn_timing.csv") %>% 
  mutate(date = as.Date(date), year = year(date), yday = yday(date))

index_sk <- spawners %>% 
  mutate(duplicate = duplicated(.)) %>% 
  filter(duplicate==FALSE) %>% 
  group_by(location, year, yday, date) %>% 
  filter(location == "Index") %>% 
  filter(species == "sk") %>% 
  summarise(
    live = if (all(is.na(live))) NA_real_ else sum(live, na.rm = TRUE),
    dead = if (all(is.na(dead))) NA_real_ else sum(dead, na.rm = TRUE)) %>% 
  arrange(date) %>% 
  ungroup() %>% 
  mutate(day_ind = as.numeric(factor(yday)), 
         year_ind = as.numeric(factor(year))) %>% 
  filter(!is.na(live)) 

index_sk %>% 
  ggplot(aes(x = yday, y = 1e-3*live, color = location))+
  geom_point(size=0.5, color='black')+
  facet_wrap(~year,scales="free_y") + 
  theme(strip.background = element_rect(color="NA", fill="NA"))+
  theme(panel.grid.minor = element_blank(), legend.position = "none")+
  labs(y="Live Spawners Present ('000)", x="Day of the Year")
```

```{r trad_AUC, include = FALSE}
trad_AUC <- index_sk %>%            
  group_by(year) %>%   
  mutate(tdiff = date - lag(date),                      
         tdiff = replace(tdiff, which(tdiff < 0), NA),
         xbar = (live + lag(live))/2, 
         fishdays = case_when(                          
           is.na(xbar) ~ live * 11/2,         
           !is.na(xbar) ~ as.numeric(tdiff) * xbar
         ),
         cumulative = cumsum(fishdays)) %>%  
  group_by(year) %>%     
  summarise(total_auc = max(cumulative, na.rm = TRUE) +
              (live[which.max(date)]*(11/2)), #this adds in the right tail.
            nerkids = round(total_auc/11,0))
```

```{r stan, include = FALSE}
day_stand <- min(index_sk$yday)

priors <- data.frame(prior = c("log_runs_mu", "arrival_mu", "arrival_sigma", "spread","spread_sigma","residence", "count_dispersion"),
                     v1 = c(10,280-day_stand, 0,log(8-1), 0,log(11-1), 1), 
                     v2 = c(1,2,1,0.2, 10, 0.3, 0.2))


sp_dat = list(n_priors = nrow(priors), 
              priors = data.matrix(priors[,-1]),
              n_years = max(index_sk$year_ind),
              year = as.numeric(factor(index_sk$year)),
              day = index_sk$yday-day_stand,
              n_obs = nrow(index_sk),
              live_counts = index_sk$live)

init_fun <- function() list(
      log_run = rnorm(sp_dat$n_years, 10, 1),
      arrival_mu = rnorm(1, 40, 2),
      arrival_sigma = rexp(1, 1), 
      arrival_spread_mu = rnorm(1, log(7), 0.2),
      exit_spread_mu = rnorm(1, log(7), 0.2),
      exit_lag_mu = rnorm(1, log(10), 0.3)
    )


run_or_load_stan <- function(model_name, stan_path, data_list, chains = 4, cores = 4) {
  fit_file <- paste0("../model_fits/", model_name, ".RData")
  hash_file <- paste0("../model_fits/", model_name, ".md5")
  current_hash <- unname(tools::md5sum(stan_path))
  
  if (!file.exists(fit_file) || !file.exists(hash_file) || readLines(hash_file) != current_hash) {
    message("Fitting ", model_name, "...")
    m <- stan_model(file = stan_path)
    fit <- sampling(m, data = data_list, chains = chains, cores = cores, control = list(adapt_delta = 0.95), init = replicate(4, init_fun(), simplify = FALSE))
    save(fit, m, file = fit_file)
    writeLines(current_hash, hash_file)
  } else {
    message("Loading saved fit for ", model_name, "...")
    load(fit_file)
  }
  return(fit)
}

if(!file.exists("../stan/spawners_m1.stan")) stop('STOP no file ../stan/spawners_m1.stan')
if(!dir.exists("../model_fits")) stop('STOP no folder ../model_fits')

fit_m1 <- run_or_load_stan("spawner_m1", "../stan/spawners_m1.stan", sp_dat)
fit_m2 <- run_or_load_stan("spawner_m2", "../stan/spawners_m2.stan", sp_dat)
fit_m3 <- run_or_load_stan("spawner_m3", "../stan/spawners_m3.stan", sp_dat)
fit_m4 <- run_or_load_stan("spawner_m4", "../stan/spawners_m4.stan", sp_dat)
fit_m5 <- run_or_load_stan("spawner_m5", "../stan/spawners_m5.stan", sp_dat)
fit_m6 <- run_or_load_stan("spawner_m6", "../stan/spawners_m6.stan", sp_dat)
fit_m7 <- run_or_load_stan("spawner_m7", "../stan/spawners_m7.stan", sp_dat)
fit_m8 <- run_or_load_stan("spawner_m8", "../stan/spawners_m8.stan", sp_dat)
fit_m8_MVN <- run_or_load_stan("spawner_m8_MVN", "../stan/spawners_m8_MVN.stan", sp_dat)

# model checks
worst_Rhat <- summary(fit_m8_MVN)$summary %>% 
  as.data.frame() %>% 
  mutate(Rhat = round(Rhat, 3)) %>% 
  arrange(desc(Rhat))

worst_Rhat %>% 
  #filter(n_eff>3) %>% 
  ggplot(aes(x = n_eff, y = Rhat))+
  geom_point()+
  geom_hline(yintercept = 1.01, lty = 2)+
  geom_vline(xintercept = 400, lty = 2)

traceplot(fit_m8_MVN, pars = rownames(worst_Rhat)[1:9])

loo_m1 <- loo(fit_m1)
loo_m2 <- loo(fit_m2)
loo_m3 <- loo(fit_m3)
loo_m4 <- loo(fit_m4)
loo_m5 <- loo(fit_m5)
loo_m6 <- loo(fit_m6)
loo_m7 <- loo(fit_m7)
loo_m8 <- loo(fit_m8)
loo_m8_MVN <- loo(fit_m8_MVN)

loos <- list(
  m1 = loo_m1,
  m2 = loo_m2,
  m3 = loo_m3,
  m4 = loo_m4,
  m5 = loo_m5,
  m6 = loo_m6,
  m7 = loo_m7,
  m8 = loo_m8,
  m9 = loo_m8_MVN)

cmp <- loo_compare(loos)

loo_df <- as.data.frame(cmp) %>%
  tibble::rownames_to_column("model") %>%
  rename(
    dloo = elpd_diff,
    se = se_diff
  ) |> 
  arrange(model) |> 
  mutate(model = as.numeric(stringr::str_extract(model, "\\d+")))


fit <- fit_m8_MVN
post <- rstan::extract(fit)
```

# Introduction

## Spawner survey analyses

Functions relating the abundance of spawners ("stock") to the subsequent
abundance of their offspring returning as adults ("recruits" to
fisheries) are fundamental to the management of salmon fisheries and
salmon habitats [@ricker1975; @hilborn1992]. For salmon, spawner
abundance in a river, stream, or lake beach is typically estimated
through repeated visual surveys – from a boat, a helicopter, a drone,
while snorkeling or walking – that count the number of spawners present
across multiple days during the spawning period. As part of the survey
process, counts are adjusted for observer efficiency (percent observed
of salmon present) @hilborn1999 and extrapolated to survey area not
observed. Those adjustments are not considered in what follows.
Converting repeated daily estimates of spawners *present* into a yearly
estimate of spawner *abundance* requires accounting for two factors: (1)
surveys typically do not cover every day of the spawning period, and (2)
spawning fish generally remain on the spawning grounds for multiple
days: 8.8-18.7 days for sockeye salmon; @perrin1990. Conventional
methods adjust for these factors but rely on assumptions that can
introduce bias and fail to account for uncertainty.

The classic, and still widely used method, is based on *trapezoidal area
under the curve* [TAUC\; @ames1977; @perrin1990]. This is a
non-parametric approach, where "curve" means linear interpolations of
counts between survey dates ([a Reimann
sum](https://en.wikipedia.org/wiki/Riemann_sum "wikipedia")) to fill in
missing survey days [@english1992; @irvine1992]. Prior knowledge about
first and last dates for presence of spawners is required and often
assumed. The resulting integral has units of $fish \times days$. To
account for repeat observations of individual fish, total spawner
abundance is calculated by dividing $fish \times days$ by the average
**residence time** (survey life) of a spawner on the surveyed spawning
ground. Residence time has been found to vary across years and across
spawning grounds, so it is recommended that annual, spawning-ground
specific estimates should be used [@bocking1988; @perrin1990;
@irvine1992]. There are a range of methods for estimating residence time
[reviewed in @perrin1990]. Estimating residence time is difficult and
requires additional surveying, so residence time is often assumed,
sometimes based on similarity between populations [e.g., @hyatt2010].
Residence time acts as a scalar on the survey counts and can introduce
substantial bias to spawner abundance estimates if the assumed value is
incorrect.

To account for observation error, a continuous curve (Gaussian normal if
symmetric, else beta @hilborn1999 or Weibull @labelle2021 as examples)
can be fit to the time series of survey estimates in a year using linear
regression [@millar2012]. This produces a *Gaussian area under curve*
(GAUC), where total spawning abundance is the integral of the curve,
corrected for residence time as for TAUC. This parametric approach has
been extended by applying hierarchical regression to analyze multiple
years or spawning grounds simultaneously, enabling more reliable
estimates of spawner abundance when data are sparse by leveraging
information shared across years or locations [@adkison2001; @su2001].
GAUC improves upon TAUC by estimating precision for spawner abundance,
but still relies on an assumed (independently estimated) residence time.
Additionally, it assumes that the arrival and exit processes -- due to
death for semelparous salmon -- combine to produce the assumed temporal
distribution of spawners present on the spawning grounds.

## Arrival-exit models

Mechanistic models of the arrival and exit processes @szerlong2008
address some of the limitations of the TAUC and GAUC approaches. As
initially developed for salmon by @quinn1997, the number of spawners
present during a survey is arrivals cumulative to a survey date, minus
exits cumulative to that date (@fig-enterexit). Residence time, as the
lag between mean arrival date and mean exit date, is then an estimated
parameter rather than an assumption. In the subsequent model by
@hilborn1999, arrivals and exits were assumed to follow the same
cumulative distribution. @su2001 and @adkison2001 extended this
framework to make residence time a declining function of arrival date.
They also allowed for year-to-year variation in the timing parameters,
estimated for all years simultaneously in hierarchical regression to
improve estimates in years with sparse data. Further variations were
explored by @labelle2021, including multiple modes of spawner arrivals.
Despite its practical utility (@holt2008a), the hierarchical
arrival-exit process model has not been widely adopted to estimate
spawner abundance.

This paper introduces further evolution of the arrival-exit approach
[e.g., @hilborn1999; @su2001]. Our model differs from preceding by
estimating four timing parameters: (1) **arrival timing**, the average
day of spawner arrivals; (2) **arrival spread**, the variation (as
standard deviation, assuming Gaussian) in spawner arrival dates; (3)
**exit lag**, the difference between mean arrival date and mean exit
date; and (4) **exit spread**: the variation in exit timing
(@fig-enterexit). This avoids imposing a specific function relating
arrival date and and residence time, as in @su2001. As a further
improvement, we model the timing parameters as a multivariate normal
distribution (MVN) with an estimated covariance structure. This
recognizes the likelihood of correlations among timing parameters, by
quantifying and applying these. We use a hierarchical Bayesian
framework, enabling estimation of year-to-year variation in both spawner
timing and abundance. This hierarchical structure shares information
across years, improving estimates when data are sparse while reducing
the risk of over-fitting.

We first demonstrate the performance of the model by showing how it can
accurately estimate spawner timing and abundance from simulated data.
Then we apply it to 25 years of visual spawner enumeration surveys in
the Okanagan River, to demonstrate its utility for ecological inference
and for management-relevant estimation of spawner abundance.

```{r fig-enterexit, echo=FALSE, warning=FALSE, message=FALSE, fig.height = 4, fig.width = 6, fig.align= 'center', fig.cap = "Example of how the model estimates total spawner abundance, arrival timing, arrival spread, exit lag, and exit spread based on the observed counts (points) of spawners on multiple days. The number of spawners present on a given day is the difference between accumulated arrivals and accumulated exits. We assumed spawner counts are imperfect observations of the true number (sampling error), so individual points vary from the estimated presence. The abundance of spawners is estimated as the total number that arrived. All model parameters are estimated with uncertainty – lines show the posterior median, bands show the 95% credible intervals. For the arrival and exit curves, uncertainty grows as abundance increases. This is because uncertainty in the timing of arrivals and exits combines with the uncertainty in the overall abundance estimates."}
j <- 25 # used 2024 as example
arrived <- sapply(15:max(sp_dat$day), FUN = function(x) {
  pnorm(x, post$arrival[,j], post$arrival_spread[,j]) * exp(post$log_run[,j])
}) %>% 
  as_tibble() %>%
  mutate(draw = row_number()) %>%
  pivot_longer(-draw, names_to = "day", values_to = "spawners") %>%
  mutate(day = as.integer(gsub("V", "", day)) + 14) %>% 
  mutate(type = "arrived")

exited <- sapply(15:max(sp_dat$day), FUN = function(x) {
  pnorm(x, post$arrival[,j] + post$exit_lag[,j], post$exit_spread[,j]) * exp(post$log_run[,j])
}) %>% 
  as_tibble() %>%
  mutate(draw = row_number()) %>%
  pivot_longer(-draw, names_to = "day", values_to = "spawners") %>%
  mutate(day = as.integer(gsub("V", "", day))+14) %>% 
  mutate(type = "exited")

counts <- sapply(15:max(sp_dat$day), FUN = function(x) {
  entered <- pnorm(x, post$arrival[,j], post$arrival_spread[,j])
  exited <- pnorm(x, post$arrival[,j] + post$exit_lag[,j], post$exit_spread[,j])
  
  exp(post$log_run[,j] + log(entered - (entered * exited)))
}) %>% 
  as_tibble() %>%
  mutate(draw = row_number()) %>%
  pivot_longer(-draw, names_to = "day", values_to = "spawners") %>%
  mutate(day = as.integer(gsub("V", "", day)) + 14) %>% 
  mutate(type = "present")

example_data <- bind_rows(arrived, exited, counts) %>% 
  group_by(type, day) %>% 
  summarise(l89 = quantile(spawners, probs = 0.065), 
            u89 = quantile(spawners, probs = 0.945),
            spawners = median(spawners))

mean_arrival_day <- median(post$arrival[,j])
median_residence <- median(post$exit_lag[,j])
y_value <- median(pnorm(mean_arrival_day, post$arrival[,j], post$arrival_spread[,j]) * exp(post$log_run[,j]))
x_start <- mean_arrival_day + day_stand
x_end <- mean_arrival_day + median_residence + day_stand

residence_df <- data.frame(
  x = x_start,
  xend = x_end,
  y = y_value,
  yend = y_value,
  label = "residence"
)

ggplot(example_data, aes(x = day + day_stand, y = spawners, group = type))+
  geom_ribbon(aes(ymin = l89, ymax = u89), color = NA, alpha = 0.2)+
  geom_textline(size = 4, linewidth = 1, aes(label = type), hjust = 0.65, family = "sans") +
  scale_x_continuous(labels = ~ format(as.Date(.x, origin = "2023-12-31"), "%b ") %>% 
                       paste0(as.integer(format(as.Date(.x, origin = "2023-12-31"), "%d"))),
                     breaks = as.numeric(as.Date(c("2024-09-15", "2024-10-01", "2024-10-15", "2024-11-01", "2024-11-15")) - as.Date("2023-12-31")), 
                     name = "", expand = c(0, 0))+
  scale_y_continuous(labels = label_number(scale = 1e-3, suffix = "k"), name = "Spawners", expand = expansion(mult = c(0, 0)), limits = c(0, 1.3e5), breaks = seq(0, 1.25e5, by = 0.25e5))+
  geom_point(data = index_sk %>% filter(year == 2024), aes(x = yday, y = live, group = NA))+
    #geom_textsegment(data = residence_df, aes(x = x_start, xend = x_end, y = y_value, yend = y_value, label = "residence"), linewidth = 0.5, inherit.aes = FALSE, size = 2.8, family = "sans")+
  annotate("text",
           x = Inf,
           y = max(example_data$spawners),  
           label = "total abundance",
           hjust = 1.03,
           vjust = -0.3,
           size = 4,
           family = "sans")+
  theme_bw()+
  theme(panel.grid = element_blank())

```

# Methods

## Model

We modelled spawner abundance based on observed spawner counts using a
series of hierarchical Bayesian models fit in the Stan programming
language [@standevelopmentteam2020], using the rStan package
[@standevelopmentteam2020] in R v4.4.0 [@rcoreteam2025]. These models
varied in the degree to which they allowed for interannual variation in
the four timing parameters (arrival timing and spread, exit lag and
spread) and whether they were jointly estimated using a multivariate
normal distribution (MVN) to account for correlations among the timing
parameters. All models follow the processes shown in @fig-enterexit.
These models only deal with live spawners. Spawner counts are integers
and can include zero, therefore we used a negative binomial distribution
for the observation model:

$$N_{y,d} \sim \text{NegBinomial}(\mu_{y,d}, \phi)$$ {#eq-obs} where
$N_{y,d}$ is the PDD for predicted counts, $\mu_{y,d}$ is the expected
number of live spawners present in year $y$ on day $d$, and $\phi$
captures variability around $\mu_{y,d}$, including both observation
error and natural variation that is not explicitly modelled. We assume
that this expected number of spawners present, $\mu_{y,d}$ is a function
of three unobserved (i.e., latent) quantities that are estimated in our
model: (1) $p^\text{arrived}_{y,d}$, the proportion of spawners that
have arrived by day $d$; (2) $p^\text{exited}_{y,d}$, the proportion of
spawners that have exited by day $d$, (*i.e.*, died or left the spawning
grounds), and (3) $N_y$, the total spawner abundance in that year:
$$\mu_{y,d} \approx p^\text{arrived}_{y,d} \times (1 - p^\text{exited}_{y,d}) \times N_y.$$ {#eq-live}

We assume that the spawners arrive over a range of days that follows a
Gaussian distribution, reside there for some days of spawning activity,
and then die or leave the spawning grounds. Other distributions for
skewed or multimodal arrivals may be applicable. Thus, the proportion of
spawners that have arrived by a given date $d$ in year $y$ follows a
cumulative normal distribution:
$$p^\text{arrived}_{y,d} = \Phi\left( \frac{ d - \mu^\text{arrival}_y }{ \sigma^\text{arrival}_y } \right).$$ {#eq-arrived}

Likewise, the proportion of arrived spawners that have exited the live
count by a given date $d$ in year $y$ follows a second cumulative normal
distribution:
$$p^\text{exited}_{y,d} = \Phi\left( \frac{ d - (\mu^\text{arrival}_y + \mu^\text{exit lag}_y)}{\sigma^\text{exit}_y} \right),$$ {#eq-exited}
where $\mu^\text{exit lag}_y$ defines the number of days by which the
mean exit date lags behind the mean arrival date $\mu^\text{arrival}_y$.
Total spawner abundance, $N_y$, in each year is a primary variable of
interest and is estimated by the model as the scaling factor that links
the estimated arrival and exit processes to the expected live spawners,
$mu_{y,d}$. We set a prior on spawner abundance as
$log(N_y) \sim \text{Normal}$(`r priors[1,2]`, `r priors[1,3]`). This
weakly informative prior is consistent with the range of log-transformed
spawner abundances previously estimated for this population using
trapezoidal AUC, but does not constrain estimates to match those values.

To prevent numerical instability when equation @eq-live approaches zero,
we use a log-transformed approximation of the same form:
$$\mu_{y,d} =  \exp\left( \log\left(1 + 10^4 \times p^\text{arrived}_{y,d} \times (1 - p^\text{exited}_{y,d})\right) - \log(10^4) \right) \times N_y$$ {#eq-livefull}
where the factor $10^4$ serves as a scaling factor that prevents changes
in near-zero values from having an abrupt influence on $\mu_{y,d}$ .

In the equations above, we specified $\mu^\text{arrival}_y$,
$\sigma^\text{arrival}_y$, $\mu_y^\text{exit lag}$ and
$\sigma^\text{exit}_y$ using subscript $y$ to indicate that they vary
from year to year. Some of these timing parameters may not need to vary
annually in a model applied to a specific situation, so we compared a
series of models with different combinations of varying and constant
versions of these parameters (summarized in @fig-modcompare A) and
evaluated their fit to withheld data (Leave One Out, @vehtari2017a), as
described in section 2.1.1.

The time-varying versions of these parameters were estimated using a
hierarchical structure, where yearly values were drawn from a
hyperdistribution (e.g.,
$\mu^\text{arrival}_y \sim \text{Normal}(\mu^\text{arrival}_\mu, \mu^\text{arrival}_\sigma)$)
defined by the multiyear mean and year-to-year standard deviation. We
set a prior for the arrival day group mean as
$\mu^\text{arrival}_\mu \sim \text{Normal}$(`r priors[2,2]`,
`r priors[2,3]`), corresponding to a mean of approximately
`r round(priors[2,2]/max(sp_dat$day)*100)`% of the typical spawning
survey period, with a prior interval spanning
`r round((priors[2,2] - 1.96*priors[2,3])/max(sp_dat$day)*100)`--`r round((priors[2,2] + 1.96*priors[2,3])/max(sp_dat$day)*100)`
days. This reflects our expectation that peak arrivals occur during the
early to mid portion of the survey window. We used the prior
$\mu_\sigma^{\text{arrival}} \sim \text{Exponential}$(`r priors[3,3]`)
to represent a moderate level of expected year-to-year variation in
$\mu_y^{\text{arrival}}$. We modelled $\sigma^\text{arrival}_y$,
$\mu_y^\text{exit lag}$ and $\sigma^\text{exit}_y$ using priors of the
form $\text{Exponential}(\lambda)+1$ to ensure that parameters were
strictly positive and bounded away from zero following the biological
constraint that spawners vary in their arrival and exit timing, and that
they remain on the spawning grounds for at least one day. For these
spread and lag parameters, the hierarchical structure is applied on the
log-transformed scale to improve sampling and ensure positivity. We set
priors for $\sigma_\mu^\text{arrival}$ and $\sigma_\mu^\text{exit}$ as
$\sim \text{Normal}(\log$(`r exp(priors[4,2])`), `r priors[4,3]`),
corresponding to an expectation that arrival and exit timing should have
similar spread. This corresponds to an expected range of duration
between `r round(exp(priors[4,2] - 1.96* priors[4,3])+1)` and
`r round(exp(priors[4,2] + 1.96* priors[4,3])+1)` days. We set the prior
for $\mu_\mu^\text{exit lag}$ as
$\sim \text{Normal}(\log$(`r exp(priors[6,2])`), `r priors[6,3]`),
aligning with the previously assumed residence time of approximately 11
days, and with likely values ranging from
`r round(exp(priors[6,2] - 1.96* priors[6,3])+1)` to
`r round(exp(priors[6,2] + 1.96* priors[6,3])+1)` days. We used the same
$\sim \text{Exponential}$(`r priors[5,3]`) prior for
$\mu_\sigma^{\text{spread}}, \mu_\sigma^{\text{exit}}$, and
$\mu_\sigma^{\text{exit lag}}$ to represent a moderate level of expected
year-to-year variation in these parameters on the log-transformed scale.


In the models where we use constant (not varying by year) versions of
the timing parameters, we drop the subscript $y$ and denote then as:
$\mu^\text{arrival}$, $\sigma^\text{arrival}$, $\mu^\text{exit lag}$,
and $\sigma^\text{exit}$. In these cases, we use the prior for the group
level $\mu$ and omit the group-level variance parameter, thus removing
the hierarchical structure.

In the model where all timing parameters have interannual variation, we
compare a version where they are estimated independently to a model
where they are jointly estimated using a multivariate normal
distribution (MVN). We specify this joint estimation as:
$$\boldsymbol{\theta_y} \sim \text{MVN}(\boldsymbol{\mu}, \boldsymbol{\Sigma}),$$
where $\boldsymbol{\theta_y}$ is a vector of the four timing parameters
in each year,

$$\boldsymbol{\theta_y} = [\mu^\text{arrival}_y, \log(\sigma^\text{arrival}_y - 1), \log(\sigma^\text{exit}_y - 1), \log(\mu^\text{exit lag}_y - 1)],$$
and $\boldsymbol{\mu}$ is a vector of the timing parameter group level
means,
$$\boldsymbol{\mu} = [\mu^\text{arrival}_\mu, \log(\sigma^\text{arrival}_\mu - 1), \log(\sigma^\text{exit}_\mu - 1), \log(\mu^\text{exit lag}_\mu - 1)].$$
The $\log(x - 1)$ transformation for the spread and lag parameters
ensures that they remain positive and bounded away from zero, while also
improving sampling stability, as described above. The parameter
$\boldsymbol{\Sigma}$ is the residual covariance matrix, which describes
both the amount of year-to-year variation in each timing parameter and
correlation among them. This structure jointly estimates the timing
processes, allowing the model to capture interdependence between them,
for example, if exit lag is longer in years when arrival timing is
earlier.

We used ChatGPT (model 4o) as a tool to support Stan model development,
including brainstorming model structure and identifying causes of
convergence issues. ChatGPT was not used to generate Stan code directly,
and all suggestions were reviewed and verified by the authors for logic
and accuracy.

### Diagnostics

An essential concept in Bayesian statistics is that the data are known
but the best model for those data is unknown [@mcelreath2020] and always
a vast simplification of reality [@jaynes2003]. Metrics to compare and
rank competing models are required. In ecological systems, a
superabundance of habitat indicators as covariates means over-fitting is
a problem. Leave-one-out (LOO) statistics\[\^1\], where each point is
predicted while omitted from fitting (cross-validation), provides
diagnostics for over-fitting. We used ELPD, the sum of *expected log
predictive probability density* for each of $n$ omitted observation's
prediction, and $\mathrm{ELPD}_{SE}=\text{sd}(\mathrm{ELPD})/\sqrt{n}$
the standard error of those probabilities [@vehtari2017].

### Mean residence

Although $\mu^\text{exit lag}_y$ defines the average delay between
arrival and exit distributions, it does not equal the mean residence
time (as used by TAUC). This is because residence time depends on the
interactions of arrival and exit processes and is not directly estimated
by any single parameter of the model. Importantly, whenever
$\sigma^\text{arrival}_y \neq \sigma^\text{exit}_y$, residence time
cannot be the same for all individuals, and instead varies depending on
when each spawner arrives. We can, however, calculate a mean residence
time in each year based on the sum of estimated number of live spawners
across all days in the spawning period, $D$, divided by the estimated
total spawner abundance:
$$\mu^{\text{residence}}_y = \frac{\sum_{d = 1}^D \left(p^\text{arrived}_{y,d} \times (1 - p^\text{exited}_{y,d}) \times N_y^{\text{spawners}}\right)}{N_y^{\text{spawners}}}.$$ {#eq-residence}
This is the equivalent of back calculating mean residence time from the
area under the curve, which is possible because we have directly
estimated $N_y^{\text{spawners}}$. However, because
$N_y^{\text{spawners}}$ appears in both the numerator and the
denominator of @eq-residence, we can simplify this equation as:
$$\mu^{\text{residence}}_y = \sum_{d = 1}^D \left(p^\text{arrived}_{y,d} \times (1 - p^\text{exited}_{y,d})\right).$$ {#eq-residence-simple}

## Simulations
```{r simulations, include = FALSE}
#values that define simulations
log_run_mu               <-  10    # log value for lognormal PDD, 
log_run_sigma            <-   1    # log value for lognormal PDD, 
arrival_timing_mu        <-  35    # normal  
arrival_timing_sigma     <-   1    # 95% days 33 to 37 
arrival_spread_mu        <-   4    # normal
arrival_spread_sigma     <-   0.2  # 95% 3.6 to 4.4
exit_lag_mu              <-  10    # normal 
exit_lag_sigma           <-   0.5  # 95% 9 to 11
exit_spread_mu           <-   3    # normal. should be < arrival_spread_mu
exit_spread_sigma        <-   0.2  # 95%  2.6 to 3.4
years_n                  <-  20    # number of years, 4x3 plot
days_n                   <- 100    # initial number of days, checked below.
samp_accuracy            <-   0.15 # rnorm(n, 1, 0.15) 
# 15% sampling error, +/- 95% = 0.3 is a lot. Likely under not over. 
gaps                     <- c(1,3,5,7) # days between surveys. survey_gaps identifies scenarios.

sim_params <- data.frame(parameter = c("log run", "arrival", "arrival spread", "exit lag", "exit spread"),
           mu = c(log_run_mu, arrival_timing_mu, arrival_spread_mu, exit_lag_mu, exit_spread_mu),
           sigma = c(log_run_sigma, arrival_timing_sigma, arrival_spread_sigma, exit_lag_sigma, exit_spread_sigma),
           distribution = c("Normal"))

if(!file.exists("../data/simulations.RData")){
m <- stan_model("../stan/spawners_m8.stan")


# derived
days  = 1:days_n                   # overrides lubridate::days()
years = 1:years_n                  # overrides lubridate::years()
sim = c(arrival_timing_mu, arrival_timing_sigma,
        arrival_spread_mu, arrival_spread_sigma,
        exit_spread_mu, exit_spread_sigma, 
        exit_lag_mu, exit_lag_sigma)


n_sim <- 20
sim_summary <- data.frame()
s <- 1
while(s <=n_sim){
  print(paste0("sim", s))
  #single simulation
  draws = data.frame(  year = years,
                       run             = rlnorm(years_n,  log_run_mu, 
                                                log_run_sigma) %>% round,
                       arrival_timing  = rnorm(years_n, arrival_timing_mu,
                                               arrival_timing_sigma) %>% round(2),
                       arrival_spread  = rnorm(years_n,arrival_spread_mu,
                                               arrival_spread_sigma)  %>% round(2),
                       exit_lag        = rnorm(years_n, exit_lag_mu, 
                                               exit_lag_sigma) %>% round(2),
                       exit_spread     = rnorm(years_n,exit_spread_mu, 
                                               exit_spread_sigma) %>% round(2) )
  
  sim_values.df <- draws |> 
    rename(arrival = arrival_timing) |> 
    mutate(log_run = log(run)) |> 
    gather(key = .variable, value = true_value, -year)
  
  present <- sapply(1:years_n, function(i){
    entered <- pnorm(days, draws$arrival_timing[i],  draws$arrival_spread[i])
    exited  <- pnorm(days, draws$arrival_timing[i] + draws$exit_lag[i], draws$exit_spread[i])
    present <- (entered * (1-exited) * draws$run[i]) # zero if out before in
  }) 
  
  dat_sim = data.frame(
    year    = rep(years,each=days_n),
    day     = rep(days,years_n),
    present = present%>% as.vector %>% round)
  # first and last days of presence over all year.
  a= dat_sim$day[dat_sim$present > 0]
  day_first = min(a)
  day_last  = max(a) 
  
  dat_sim %<>% mutate(sampled = present*rnorm(nrow(dat_sim),1, samp_accuracy))
  dat_sim %<>% mutate(sampled = floor(sampled)) # truncate: 0.9 fish -> 0 fish.
  # determine survey dates with 1, 3, 5, 7 day gaps. 
  # save four survey scenarios. Same dates all years
  days_surveyed <- list(
    seq(day_first,day_last, by=2), 
    seq(day_first,day_last, by=4), 
    seq(day_first,day_last, by=6), 
    seq(day_first,day_last, by=8) )
  # create long form data frame for scenarios
  sampled.df <- data.frame()
  for(i in 1:length(days_surveyed)){
    sampled.df <- bind_rows(sampled.df, 
                            dat_sim %>% filter(day %in% days_surveyed[[i]]) %>%
                              mutate(survey_gaps=gaps[i] ))
  }
  # delete if present =< 0 or sampled =< 0 
  sampled.df %<>% filter((sampled > 0) & (present > 0))
  
  sp_dat_0 = list(
    n_priors     = nrow(priors),
    priors      = data.matrix(priors[,-1]),
    n_years     = years_n, 
    arrival_mu  = arrival_timing_mu )
  
  annual_medians= data.frame() # fitted medians annual parameters
  # loop through the four scenarios of survey gaps
  sim_fail <- FALSE         
  for (j in gaps){   # 1,3,5,7 
    # insert scenario data into stan data  
    k = sampled.df$survey_gaps == j  # which rows
    sp_dat = list_modify(sp_dat_0,
                         n_obs = sum(k),
                         year  = sampled.df$year[k],
                         day   = sampled.df$day[k],
                         live_counts= sampled.df$sampled[k] %>% as.integer )
    
    fit <- sampling(m, data = sp_dat,
                    chains = 4, cores = 4,control = list(adapt_delta = 0.95), init = replicate(4, init_fun(), simplify = FALSE))
    Rhat_max <- max(summary(fit)$summary[,"Rhat"], na.rm = TRUE)
    n_eff_min <- min(summary(fit)$summary[,"n_eff"], na.rm = TRUE)
    div_trans <- sum(sapply(get_sampler_params(fit, inc_warmup = FALSE), function(x) sum(x[, "divergent__"])))
    
    if (Rhat_max > 1.01 | div_trans > 0) {
      sim_fail <- TRUE
      break     # stop checking other gaps for this sim
    }  
    annual_medians <- bind_rows(annual_medians, 
                                spread_draws(
    fit,
    log_run[year],
    arrival[year],
    arrival_spread[year],
    exit_spread[year],
    exit_lag[year]
  ) %>%
  gather(key = .variable, value = .value,
         log_run, arrival, arrival_spread, exit_spread, exit_lag) %>%
  group_by(year, .variable) %>%
  summarise(
    est_value = median(.value, na.rm = TRUE),
    l95       = quantile(.value, 0.025, na.rm = TRUE),
    u95       = quantile(.value, 0.975, na.rm = TRUE),
    precision = median(abs(.value - median(.value, na.rm = TRUE)), na.rm = TRUE),
    .groups   = "drop"
  ) %>%
  mutate(
    gap       = j,
    Rhat_max  = Rhat_max,
    n_eff_min = n_eff_min,
    div_trans = div_trans
  )
  )
  }
  
  if (!sim_fail) {
    sim_summary1 <- left_join(annual_medians, sim_values.df) |> mutate(diff = true_value - est_value, 
                                                                       in_95 = u95 > true_value & true_value >l95, 
                                                                       ratio = est_value/true_value) |> 
      arrange(.variable) |> 
      mutate(sim = s)
    sim_summary <- bind_rows(sim_summary, sim_summary1)
  }
  s <- s+1
}

save(sim_summary, file = "../data/simulations.RData")
} else{
  load("../data/simulations.RData")
}
```

To quantify how reliably the statistical model could estimate the annual values of run size and the four timing parameters, we challenged our model with simulated data. The simulations were generated using the same parameters as the statistical model. Each simulation consisted of 20 years of observations, where the annual parameters governing the spawning timing and run size were drawn randomly from the distributions in @tbl-sim. These yearly parameters are considered to be the *true* values to which we compare our estimated values from the statistical model. The resulting *true* data was modified by multiplicative normal sampling error (15% accuracy) to generate *observed* data. To test the sensitivity of our model to different sampling frequency, we contrasted results when we subsampled these timeseries with 1, 3, 5 and 7 day gaps. We then ran 20 replicate simulations to ensure that our results were representative. We fit these observed data with statistical model 8 (@fig-modcompare), which is the same as our selected model 9 but without the multivariate correlation structure between parameters. Because our goal was to test recovery of known values rather than parameter correlations, we did not include this additional complexity in our simulations. 

For each year, we quantified the median absolute deviation (MAD, a robust version of standard deviation) of the posterior draws around the median and bias as the ratios of fitted over true values. We also quantified coverage, which is the proportion of years in which the true value fell within the 95% posterior credible interval.

```{r tbl-sim, echo=FALSE, warning=FALSE, message=FALSE}
#| tbl-cap: "Simulation parameters used in the study."
knitr::kable(
  sim_params,
  booktabs = TRUE,
  linesep = "",
  longtable = TRUE
) |>
  kableExtra::kable_styling(latex_options = "repeat_header")
```



## Okanagan sockeye

Our analysis of Okanagan sockeye was confined to the population of wild
Sockeye Salmon that spawn in October and November in the Okanagan River
(@fig-map) upstream of Osoyoos Lake. After emerging from spawning
gravel, mid-April to mid-May, they spend one or two summers (rarely
three) as parr in the North Basin of Osoyoos Lake before an ontological
shift (near autumnal equinox) to pre-smolts and subsequent migration the
following spring as smolts through the reservoirs (with abundant
predators), spillways, and smolt passage tunnels of ten hydroelectric
dams to reach the Pacific Ocean via the Columbia River @hyatt2019. After
one to three years (rarely more) of marine growth, adults return to the
Columbia River in late summer and attempt the transit of ten dams in a
950 km freshwater migration, before entering Osoyoos Lake and
subsequently spawning yet further upstream. Ascending the Okanagan River
from the Columbia River is increasingly compromised by high river
temperatures [@hyatt2020; @major1967; @quinn1997a].

Spawner estimates for Osoyoos sockeye have historically been based on
trapezoidal AUC, with an assumed residence time of 11 days [@hyatt2010].
This number was selected based on the estimate for the early Stuart
sockeye stock [@perrin1990] because that stock has similar average body
size, migration distance, and migratory timing [@hyatt2010]. Later,
@shardlow2011 used tilt tags on spawning sockeye in the Okanagan River
and estimated the residence time to be 14.1 (females) and 13.7 (males)
days in 2006, and 14.1 (females) and 11.9 (males) days in 2007.
Nevertheless, 11 days remains the consistently assumed value for AUC
estimates in this system.

![The spawning grounds for Sockeye Salmon population that rear as parr
in the North Basin of Osoyoos Lake are \~20 to 30 km upstream, in the
Okanagan River below McIntyre Dam at the outflow of Vaseux Lake. River
flows that affect spawners, eggs, and parr are monitored by Environment
Canada (EC) at the site indicated. Map by Braden Judson, Pacific
Biological Station.](../figures/Okanagan_to_Osoyoo_Lakes.png){#fig-map
fig-align="center" width="5in"}

![The spawning grounds for Sockeye Salmon population that rear as parr in the North Basin of Osoyoos Lake are \~20 to 30 km upstream, in the Okanagan River below McIntyre Dam at the outflow of Vaseux Lake. River flows that affect spawners, eggs, and parr are monitored by Environment Canada (EC) at the site indicated. Map by Braden Judson, Pacific Biological Station.](../figures/Okanagan_to_Osoyoo_Lakes.png){#fig-map fig-align="center" width="5in"}

## Spawner surveys

Spawner enumeration surveys have been conducted in the main sockeye
spawning area---also known as the Index section---of the Okanagan river
between Osoyoos and Vasseau lakes (@fig-map) since
`r min(index_sk$year)`. The Index section consists of the section from
Deer park to vertical drop structure 12 (VDS12), which is at the Oliver
bridge. All together, surveys have been conducted on `r nrow(index_sk)`
days, ranging between `r min(table(index_sk$year))` and
`r max(table(index_sk$year))` surveys per year. All surveys occurred
between
`r format(min(lubridate::make_date(2000, lubridate::month(index_sk$date), lubridate::day(index_sk$date))), "%B %e")`
and
`r format(max(lubridate::make_date(2000, lubridate::month(index_sk$date), lubridate::day(index_sk$date))), "%B %e")`.

Spawner enumeration survey methods are described in detail in
@machin2019. Weekly surveys during the initial migration phase are
increased to a three-day interval during the peak spawning period
(generally October-November). A three or four person crew performs
visual surveys of the Index section from a boat. Two observers focus on
counting live sockeye (holding and spawning), one on each side of the
middle of the boat, while another one or two observers count dead
sockeye and other species (Chinook, kokanee, and rainbow trout) from the
front of the boat. In addition, side channels throughout the Index
section are walked and fish are enumerated. Since 2012, enumeration of
each individual reach within the Index section is available, however, we
have used the summed counts across all reaches, because this is the
resolution of data that is available from 2000--2011. Limited spawning
for the Osoyoos population also occurs in the Okanagan river downstream
of the Index section, and in the section between Vaseaux Lake and
Okanagan Falls Dam. However, we have excluded these sections from our
analysis as including them would require additional complexity in our
model to account for potential variation in timing. It would also
require that we account for the relative abundances across river
sections, that vary through time. This extension is a logical next step,
but is not necessary for demonstrating our method, the goal of this
manuscript.

In the Okanagan River, the abundance of co-occurring spawning Kokanee
(an anadromous subspecies of *O. nerka*) have the potential to bias
abundance estimates of Sockeye (diadromous). This concern was alleviated
by (1) determining that misidentifying Kokanee as Sockeye is a concern
for Skaha Lake Sockeye spawning downstream from Okanagan Lake @fig-map
but not for Osoyoos Lake sockeye in separate section of the Okanagan
River where the size and colour of these subspecies differ sufficiently
to preclude error (pers. comm. Karylin Alex, Okanagan Nations Assembly),
and (2) our method deals simply with survey counts, not with observation
methods. Complications from co-occuring Kokanee were addressed using a
model similar to ours by @adkison2001.

Although the survey crews count dead sockeye, the dead surveys do not
provide a consistent estimate of the accumulation of dead spawners that
could be used in our model. Dead fish are enumerated but not marked or
removed and so can potentially be counted multiple times across survey
dates. However, there is a separate *dead pitch* crew that collects
biosamples multiple times over the course of the spawning period, but at
a lower frequency than the live counts are conducted. When doing so,
this crew removes the sampled fish. Therefore, the counts of dead fish
by the enumeration crew represent accumulated dead spawners, with
infrequent removal, that is not easily accounted for in a model.

# Results

## Model selection and convergence

Based on ELPD statistics (@fig-modcompare B) we selected model 9 as the best from those explored. This model was the most complex of those considered and it included year-to-year variation in spawner abundance, arrival timing, arrival spread, and exit spread, and residence time with a multivariate normal structure in their year-to-year variation.

The effective number of parameters ($\mathrm{p}_{\mathrm{loo}}$) increased when arrival timing variation was introduced (model 1 vs. 2), and also when any of arrival spread, exit spread, or residence were allowed to vary (models 3 to 9; @fig-modcompare C). However, all models with variation in any or all of these parameters (models 3 to 9) had similar numbers of effective parameters.

Convergence was good for model 9; the highest $\hat{R}$ was `r round(max(worst_Rhat$Rhat, na.rm = TRUE),3)` and `r round(mean(worst_Rhat$n_eff>400, na.rm = TRUE)*100, 1)`% of the parameters had effective sample sizes $>400$ (min = `r round(min(worst_Rhat$n_eff, na.rm = TRUE))`).

```{r fig-modcompare, echo=FALSE, , warning=FALSE, message=FALSE, fig.height = 5, fig.width = 9, fig.align= 'center', fig.cap= "Model summary and comparison. *Panel A* shows which parameters or model structure that were held constant or not included (white) versus varied or were included (black) in each of the candidate models (rows). *Panel B* shows the relative expected log predictive density ($\\Delta\\mathrm{ELPD}$) compared to the model with the highest out-of-sample predictive accuracy. *Panel C* shows the effective number of parameters ($\\mathrm{p}_\\mathrm{loo}$). The model with the highest out-of-sample predictive accuracy is highlighted in red in panels B and C. Points and error bars (B and C) represent means and standard errors. Note, that the x-axis in panel B is pseudo log transformed (inverse hyperbolic sine) to highlight differences among the best performing models."}

signed_log10_trans <- trans_new(
  name = "signed_log10",
  transform = function(x) sign(x) * log10(abs(x)),
  inverse = function(x) sign(x) * 10^abs(x)
)

data.frame(model = 1:9,
           arrival = c("constant/not included", "varies/included", "varies/included", "varies/included", "varies/included", "varies/included", "varies/included", "varies/included", "varies/included"), 
           arrival_spread = c("constant/not included", "constant/not included","varies/included", "constant/not included", "constant/not included", "varies/included", "varies/included", "varies/included", "varies/included"), 
           exit_spread = c("constant/not included", "constant/not included", "constant/not included", "varies/included", "constant/not included", "constant/not included", "varies/included", "varies/included", "varies/included"),
           residence = c("constant/not included", "constant/not included", "constant/not included", "constant/not included", "varies/included", "varies/included", "constant/not included", "varies/included", "varies/included"),
           MVN = c(rep("constant/not included", 8), "varies/included")) %>%
  gather(key = parameter, value = value, -model) %>% 
  mutate(parameter = factor(parameter, levels = c("arrival", "arrival_spread", "exit_spread", "residence", "MVN"))) |> 
  ggplot(aes(x = parameter, y = model, fill = value))+
  geom_tile(color = "grey")+
  scale_fill_manual(values = c("NA", "black"), "")+
  scale_y_reverse(breaks = 1:9)+
  theme(legend.position = "bottom")+
  coord_cartesian(expand = FALSE)+
  theme(panel.grid = element_blank(),
        axis.ticks = element_blank())+
  labs(x = "Parameter or model structure", y = "Model")+
  scale_x_discrete(
    labels = c(
      "arrival" = expression(mu[y]^arrival),
      "arrival_spread" = expression(sigma[y]^arrival),
      "exit_spread" = expression(sigma[y]^exit),
      "residence" = expression(mu[y]^"exit lag"),
      "AR" = "AR"
    )
  )+
  theme_bw()+
  theme(panel.grid = element_blank(), legend.position = "bottom")+


ggplot(loo_df, aes(x = model, y = dloo, color = dloo==0)) +
  geom_hline(yintercept = 0, linetype = 2)+
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = dloo - se, ymax = dloo + se), width = 0) +
  coord_flip() +
  scale_y_continuous(trans = pseudolog10_trans, breaks = c(-100, -10, -1))+
  scale_x_reverse(breaks = 1:9)+
  labs(
    x = "",
    y = expression(Delta*ELPD~"(rel. to best)")
  )+
  theme(panel.grid.minor.x = element_blank())+
  scale_color_manual(values = c(1, "red"), guide = "none")+
  theme_bw()+
  
  ggplot(loo_df, aes(x = model, y = p_loo, color = dloo==0)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = p_loo - se_p_loo, ymax = p_loo  + se_p_loo), width = 0) +
  coord_flip() +
  scale_x_reverse(breaks = 1:9)+
  labs(
    x = "",
    y = expression("effective parameters (" * p[loo] * ")")
  )+
  scale_color_manual(values = c(1, "red"), guide = "none")+
  theme_bw()+
  theme(panel.grid.minor.x = element_blank())+

  
  plot_annotation(tag_levels = "A")+
  plot_layout(widths = c(1, 0.5, 0.5))
```
## Model fit to simulated data
We find that the model is able to estimate the parameters used to generate our simulated data with reasonably low bias, median absolute deviation, and good 95% posterior credible interval coverage when using data with a one day gap between surveys (@fig-sims). As the gap between surveys increases, estimates become more biased, and less precise, but the degree to which they do so varies by parameter. Our estimate of total abundance, log run size, is the least sensitive to survey gap; higher survey gaps lead to a slight negative bias, but this is only `r round(median(100*(1- sim_summary$ratio[sim_summary$.variable == "log_run" & sim_summary$gap == 7])), 1)`% with a gap of 7 days between surveys, and the proportion of years for which the true log run size falls within the 95% posterior credible estimate only decreases from nearly 100% with 1 and 3 day gaps to ~ 80% with 5 and 7 day gaps. Arrival, arrival spread, and exit spread, are more sensitive to increasing survey gaps, with their deviations from the true value increasing and these deviations becoming more negative. Exit lag is the most sensitive to increased survey gaps; our simulations show that the model struggles to accurately estimate its true value even with 1 day gaps, and increasing gaps tend to result in greater and greater overestimation. 

```{r fig-sims, echo=FALSE, , warning=FALSE, message=FALSE, fig.height = 4, fig.width = 9, fig.align= 'center', fig.cap= "Median absolute deviation (A), bias (B), and 95% credible interval coverage (C) for each parameter for simulated data as a function of increasing survey gap (days). All measures compare modelled estimates with true values used to generate the simulations. Points and intervals in panels A and B show the median, 66% and 95% distribution of yearly values, pooled across 20 replicate simulations, each consisting of 20 individual years. Panel C shows the corresponding distributions of coverage across simulations."}

  ggplot(sim_summary, aes(x = gap, y = precision * 100, color = .variable))+
  geom_hline(yintercept = 0, lty = 2)+
  stat_pointinterval(position = position_dodge(width = 0.5))+
  #scale_color_okabe_ito()+
  labs(x = "Survey gap (days)",
       y =  "Median absolute deviation (%)")+
  scale_x_continuous(breaks = seq(1, 7, by = 2))+
  theme_bw()+
  theme(legend.position = "bottom")+
    
      ggplot(sim_summary , aes(x = gap, y = (1-ratio)*100, color = .variable))+
  geom_hline(yintercept = 0, lty = 2)+
  stat_pointinterval(position = position_dodge(width = 0.5))+
  #scale_color_okabe_ito()+
  labs(x = "Survey gap (days)",
       y = "Bias (%)")+
  scale_x_continuous(breaks = seq(1, 7, by = 2))+
  theme_bw()+
    
  ggplot(sim_summary |> group_by(gap, .variable, sim) |> summarise(in_95_perc = mean(in_95)*100) , aes(x = gap, y = in_95_perc, color = .variable))+
  geom_hline(yintercept = 100, lty = 2)+
  stat_pointinterval(position = position_dodge(width = 0.7))+
  #scale_color_okabe_ito(name = "")+
  labs(x = "Survey gap (days)",
       y =  "95% CI coverage (%)")+
  scale_x_continuous(breaks = seq(1, 7, by = 2))+
  theme_bw()+
  
  plot_annotation(tag_levels = "A")+
  plot_layout(guides = "collect") &
  scale_color_okabe_ito(name = "") &
  theme(legend.position = "bottom")

```

## Model fit to Okanagan sockeye counts

Our expected counts correspond well with the observed Okanagan sockeye counts (@fig-spawnercurves; @fig-obsfit). Variation between the observed and expected is quantified in our estimate of spawner count overdispersion ($\phi^\mathrm{live}$), which had a posterior median of `r round(median(post$live_phi), 1)` (95% CI: `r round(quantile(post$live_phi, 0.025), 1)`--`r round(quantile(post$live_phi, 0.975), 1)`).

```{r fig-spawnercurves, echo = FALSE, warning=FALSE, message=FALSE, fig.cap="Observed versus estimated spawners present over the spawning period in each year (panels). Points show the observed counts, the lines and shaded bands show the posterior median the 95% credible intervals. Y-axis limits vary across panels to emphasize model fit rather than variation across years.", fig.align='center', fig.height=7, fig.width=8}
spawn_curves.df <- data.frame()
for(j in 1:ncol(post$log_run)){
  live <- sapply(1:max(sp_dat$day), FUN = function(x) {
    entered <- pnorm(x, post$arrival[,j], post$arrival_spread[,j])
    exited <- pnorm(x, post$arrival[,j] + post$exit_lag[,j], post$exit_spread[,j])
    
    exp(post$log_run[,j] + log(entered - (entered * exited)))
  }
)
  spawn_curves.df <- bind_rows(spawn_curves.df, data.frame(
      year = j + min(spawners$year)-1, 
      day = 1:max(sp_dat$day)+day_stand,
      fish = apply(live, 2, median),
      l95 = apply(live, 2, quantile, probs = 0.025),
      u95 = apply(live, 2, quantile, probs = 0.975)))
}

ggplot(spawn_curves.df, aes(x = day, y = fish))+
  geom_line()+
  geom_ribbon(aes(ymin = l95, ymax = u95), color = NA, alpha = 0.2)+
  facet_wrap(~year, scales = "free_y")+
  geom_point(data = index_sk, aes(x = yday, y = live), size = 0.8)+
  scale_y_continuous(labels = label_number(scale = 1e-3, suffix = "k"), name = "Spawners present (thousands)", expand = expansion(mult = c(0, 0.02)))+
  theme_bw()+
  theme(strip.background = element_rect(color="NA", fill="NA"))+
  scale_x_continuous(labels = ~ format(as.Date(.x, origin = "2023-12-31"), "%b"), 
                     breaks = as.numeric(as.Date(c("2024-09-01", "2024-10-01", "2024-11-01")) - as.Date("2023-12-31")), name = "")+
    theme(panel.grid.minor = element_blank())
  
```

```{r effective-residence, include = FALSE}
effective_residence <- matrix(NA, nrow = length(post$log_run[,1]), ncol = ncol(post$log_run))
for(j in 1:ncol(post$log_run)){
  live <- sapply(1:max(sp_dat$day), FUN = function(x) {
    entered <- pnorm(x, post$arrival[,j], post$arrival_spread[,j])
    exited <- pnorm(x, post$arrival[,j] + post$exit_lag[,j], post$exit_spread[,j])
    
    entered * (1 - exited)
  })
  
  effective_residence[,j] <- apply(live, 1, sum)
}

colnames(effective_residence) <- unique(index_sk$year)

```

## Spawning timing

Arrival timing ($\mu_y^\text{arrival}$) varied across years (@fig-timings A), with posterior medians ranging from `r format(lubridate::ymd("2025-01-01") + round(min(apply(post$arrival, 2, median))+day_stand - 1), "%B %e")` to `r format(lubridate::ymd("2025-01-01") + round(max(apply(post$arrival, 2, median))+day_stand - 1), "%B %e")`. Arrival spread ($\sigma_y^\text{arrival}$) varied across years (@fig-timings B), with posterior medians ranging from `r round(min(apply(post$arrival_spread, 2, median)),1)` to `r round(max(apply(post$arrival_spread, 2, median)),1)` days. Exit lag ($\mu_y^\text{exit lag}$) varied across years (@fig-timings C), with posterior medians ranging from `r round(min(apply(post$exit_lag, 2, median)),1)` to `r round(max(apply(post$exit_lag, 2, median)),1)` days. Exit spread ($\sigma_y^\text{exit}$) tended to be lower than arrival spread (@fig-timings D), with posterior medians ranging from `r round(min(apply(post$exit_spread, 2, median)),1)` to `r round(max(apply(post$exit_spread, 2, median)),1)` days.

The only timing parameters that showed strong evidence of correlations in their yearly values were arrival timing ($\mu_y^\text{arrival}$) and arrival spread ($\sigma_y^\text{arrival}$), with a posterior median correlation of `r round(median(post$Omega[,1,2]),2)` and `r round(mean(post$Omega[,1,2]>0)*100)`% of samples supporting a positive relationship (@fig-cor). There was weaker support for positive correlations between exit spread ($\sigma_y^\text{exit}$) and arrival timing (median = `r round(median(post$Omega[,1,3]),2)`; % support = `r round(mean(post$Omega[,1,3]>0)*100)`) and exit lag (median = `r round(median(post$Omega[,4,3]),2)`; % support = `r round(mean(post$Omega[,4,3]>0)*100)`. A negative correlation between exit spread and arrival spread was also weakly supported (median = `r round(median(post$Omega[,2,3]),2)`; % support = `r round(mean(post$Omega[,2,3]<0)*100)`).

Together, the timing parameters result in year-to-year variation in the mean residence time (@fig-timings E). The posterior median across years is `r round(median(effective_residence), 1)` days, but it ranges from `r round(min(apply(effective_residence, 2, median)), 1)` to `r round(max(apply(effective_residence, 2, median)), 1)` days.

```{r fig-timings, echo = FALSE, warning=FALSE, message=FALSE, fig.cap="Posterior distributions of mean arrival ($\\mu_y^{\\text{arrival}}$; A), arrival spread ($\\sigma_y^{\\text{arrival}}$; B), exit spread ($\\sigma_y^{\\text{exit}}$; C), and the resulting mean residence time ($\\mu^{\\text{residence}}_y$; D). The dashed line in panel D shows the previously assumed 11 day residence time. Points and error bars show the posterior median and the 66% and 95% credible intervals.", fig.align='center', fig.height=9, fig.width=6}
spread_draws(fit, arrival[year]) %>% 
  mutate(year = year + min(spawners$year)-1) %>%
  ggplot(aes(x = year, y = arrival + day_stand))+
  stat_pointinterval()+
  xlab("")+
  scale_y_continuous(labels = ~ format(as.Date(.x, origin = "2023-12-31"), "%b %d"), 
                     breaks = seq(280, 292, by = 3), name = bquote("Mean arrival (" * mu[y]^"arrival" * ")"))+
  theme_bw()+
  
  spread_draws(fit, arrival_spread[year]) %>% 
  mutate(year = year + min(spawners$year)-1) %>% 
  ggplot(aes(x = year, y = arrival_spread))+
  stat_pointinterval()+
  scale_y_continuous(breaks = seq(0, 20, by = 3))+
  ylab(bquote("Arrival spread (" * sigma[y]^arrival * ")"))+
  xlab("")+
  theme_bw()+
  
  spread_draws(fit, exit_lag[year]) %>% 
  mutate(year = year + min(spawners$year)-1) %>% 
  ggplot(aes(x = year, y = exit_lag))+
  stat_pointinterval()+
  plot_layout(ncol = 1)+
  ylab(bquote("Exit lag (" * mu[y]^exit_lag * ")"))+
  xlab("")+
  theme_bw()+
  
  spread_draws(fit, exit_spread[year]) %>% 
  mutate(year = year + min(spawners$year)-1) %>% 
  ggplot(aes(x = year, y = exit_spread))+
  stat_pointinterval()+
  plot_layout(ncol = 1)+
  ylab(bquote("Exit spread (" * sigma[y]^exit * ")"))+
  xlab("")+
  theme_bw()+
  
  as.data.frame(effective_residence) %>% 
  mutate(draws = row_number()) %>% 
  pivot_longer(-draws, names_to = "year", values_to = "mean_residence") %>% 
  mutate(year = as.numeric(year)) %>% 
  ggplot(aes(x = year, y = mean_residence))+
  stat_pointinterval()+
  geom_hline(yintercept = 11, lty = 2)+
  scale_y_continuous(breaks = seq(1, 30, by = 3))+
  ylab(bquote("Mean residence (" * mu[y]^residence * ")"))+
  theme_bw()+
  xlab("")+
  
  plot_annotation(tag_levels = "A")
```

```{r fig-cor, echo = FALSE, warning=FALSE, message=FALSE, fig.cap="Estimated pariwise correlations between the four timing parameters. The colour shows the posterior median value, numbers indicate the proportion of the posterior with the same sign.", fig.align='center', fig.height=4, fig.width=5}

param_labels <- c(
  "1" = expression(mu[y]^arrival),
  "2" = expression(sigma[y]^arrival),
  "3" = expression(sigma[y]^exit),
  "4" = expression(mu[y]^exit_lag)
)

spread_draws(fit, Omega[parameter1, parameter2]) %>%
  filter(parameter1 < parameter2) %>% 
  group_by(parameter1, parameter2) %>% 
  summarise(
    correlation = median(Omega),
    prop_positive = mean(Omega > 0),
    prop_negative = mean(Omega < 0)
  ) %>% 
  mutate(
    prop_direction = ifelse(prop_positive > prop_negative, prop_positive, prop_negative)
  ) %>%
  ggplot(aes(x = factor(parameter1), y = factor(parameter2), fill = correlation)) +
    geom_tile() +
    geom_text(aes(label = sprintf("%.2f", prop_direction)), color = "black") +
    scale_fill_gradient2(low = "dodgerblue", high = "red", name = "Correlation") +
    scale_x_discrete(labels = param_labels) +
    scale_y_discrete(labels = param_labels) +
    xlab("") + ylab("") +
    theme_bw()
```

## Total spawner abundance

```{r AUCvsmodel, include = FALSE}

run_intervals <- spread_draws(fit, log_run[year]) %>%
  mutate(year = year + min(spawners$year) - 1) %>%
  mutate(run = exp(log_run)) %>%
  group_by(year) %>%
  summarise(lower = quantile(run, 0.17),
            upper = quantile(run, 0.83),
            median = median(run))

AUC_in_66 <- trad_AUC %>%
  left_join(run_intervals, by = "year") %>%
  mutate(in_66 = nerkids > lower & nerkids < upper, 
         lower =  median > nerkids) 

perc_diff <- spread_draws(fit, log_run[year]) %>%
  mutate(year = year + min(spawners$year) - 1) %>%
  mutate(run = exp(log_run)) %>% 
  left_join(trad_AUC) %>% 
  mutate(perc_diff = 100 * (run-nerkids)/((run + nerkids)/2)) %>% 
   group_by(year) %>% 
  summarise(perc_diff = median(perc_diff)) %>% 
  ungroup()
  
```

Estimated posterior median annual spawner abundances ($N_y^\text{spawners}$) varied from as low as `r round(min(apply(exp(post$log_run), 2, median)),-2)` in `r which(apply(exp(post$log_run), 2, median) == min(apply(exp(post$log_run), 2, median))) + min(spawners$year)-1` to as high as `r scales::comma(round(max(apply(exp(post$log_run), 2, median)),-3))` in `r which(apply(exp(post$log_run), 2, median) == max(apply(exp(post$log_run), 2, median))) + min(spawners$year)-1`. Since 2008, spawner abundance has alternated between low and high levels in a two-year cycle.

Our estimated annual abundances are relatively close to those obtained from the trapezoidal AUC estimate with an assumed 11 day residence time, with the traditional estimate falling within our 66% credible interval in `r sum(AUC_in_66$in_66)` out of `r length(AUC_in_66$in_66)` years. Across all years, our model estimates are a median of `r -(round(median(perc_diff$perc_diff), 1))`% lower than those based on TAUC (inter-annual range: `r round(min(perc_diff$perc_diff), 1)`--`r round(max(perc_diff$perc_diff), 1)`%).

```{r fig-runs, echo = FALSE, warning=FALSE, message=FALSE, fig.cap="Posterior distributions of total spawning abundance by year (black) compared to the point estimate based on traditional trapezoidal AUC with an assumed 11 day residence (red). Points and error bars show the posterior median and the 66% and 95% credible intervals.", fig.align='center', fig.height=4, fig.width=7}
spread_draws(fit, log_run[year]) %>% 
  mutate(year = year + min(spawners$year)-1) %>% 
  ggplot(aes(x = year, y = exp(log_run)))+
  stat_pointinterval(aes(color = "modelled"))+
  geom_point(data = trad_AUC, aes(x = year, y = nerkids, color = "trad. AUC"))+
  scale_y_continuous(labels = label_number(scale = 1e-3, suffix = "k"), name = "Spawner abundance (thousands)", expand = expansion(mult = c(0, 0)))+
  coord_cartesian(ylim = c(0, 2.5e5))+
  scale_color_manual(values = c(1, "red"), "")+
  xlab("")+
  theme_bw()
```

# Discussion

Conventional approaches to estimating salmon spawner abundance, such as
the TAUC method [@ames1977; @perrin1990], rely on fixed assumptions
about residence time and ignore observation error and interannual
variability in spawning dynamics. Subsequent publications have addressed
some of these limitations individually (e.g., @millar2012;
@szerlong2008; @hilborn1999; @su2001), but improvements have not been
widely adopted. Our model builds on those advances and provides a more
general modelling framework – one that jointly estimates annual run
size, arrival timing, arrival spread, exit lag, and exit spread – while
accounting for observation error and associated uncertainty. Simulation
results demonstrate that the model can recover accurate estimates of the
parameters across a range of spawning dynamics. The application to the
`r length(unique(index_sk$year))`-year time series of Okanagan sockeye
data further illustrates its utility for providing robust abundance
estimates that can support sustainable management of the population as
well as insight into its ecology.

## Model accuracy and precision
Our simulations demonstrate that the model is able to estimate parameter values with reasonable accuracy and precision, but that this is reduced as the gap between surveys in the dataset increases. Estimates of specific spawner timing parameters, particularly exit-lag, should be considered less reliable when survey frequency is not high. However, log run size, which is the primary variable of interest in spawner enumeration surveys, is relatively robust to low survey frequency.

## Residence time

Estimating residence time directly from the spawner surveys represents
an important advance over methods that rely on an assumed value. Because
residence time directly scales the estimated spawner abundance,
incorrect assumptions can lead to substantial bias---underestimating
abundances if assumed residence is too high, or overestimating if too
low. Instead of assuming a fixed residence time, our method estimates
the arrival and exit processes in each year. This idea is not new;
@hilborn1999 demonstrated how residence time could be estimated from
visual spawner data. As in @su2001 and @adkison2001, our method uses a
hierarchical structure to estimate interannual variability in residence
time, while reducing the risk of overfitting to noisy or sparse data. In
our application to Okanagan sockeye, we estimate the median *effective
residence* time to be `r round(median(effective_residence), 1)` days,
which is close to the previously assumed 11 days [@hyatt2010], but lower
than the 11.9 to 14.3 day estimates based on tilt tags in 2006 and 2007
[@shardlow2011]. However, residence time varies across years from
`r round(min(apply(effective_residence, 2, median)), 1)` to
`r round(max(apply(effective_residence, 2, median)), 1)` days,
suggesting that assuming a fixed value likely introduces bias in
specific years. Comparing our specific estimates in 2006 and 2007 to
those from @shardlow2011, we find that our estimates are still lower. In
2006, our median residence time is
`r round(apply(effective_residence, 2, median), 1)[7]` days (95% CI =
`r round(quantile(effective_residence[,7], probs = 0.025), 1)`-`r round(quantile(effective_residence[,7], probs = 0.975), 1)`)
and @shardlow2011 estimated 14.1 days for females and 13.7 days for
males. In 2007, our median residence time is
`r round(apply(effective_residence, 2, median), 1)[8]` days (95% CI =
`r round(quantile(effective_residence[,8], probs = 0.025), 1)`-`r round(quantile(effective_residence[,8], probs = 0.975), 1)`)
and @shardlow2011 estimated 14.3 days for females and 11.9 days for
males.

## Precision

Our method provides uncertainty estimates for spawner abundance and all spawning timing parameters, an important advantage over conventional methods such as TAUC that provide point estimates without uncertainty. The GAUC method [@millar2012] already provides uncertainty estimates that account for observation error, and mechanistic arrival-exit process models [@hilborn1999] can propagate uncertainty in residence time. However, the hierarchical Bayesian structure used here enables joint estimation of all parameters and ensures that uncertainty is consistently propagated throughout the model.

Uncertainty is a key piece of information for fisheries management, enabling decision-makers to assess risk and apply the precautionary approach when managing stocks [@harwood2003]. Uncertainty estimates from spawner abundance models should be propagated to subsequent life-stage models, such as those estimating parr and smolt production or adult returns from spawners. Ignoring measures of uncertainty is essentially ignoring useful information, and can result in treating imprecise estimates as overly precise, which may obscure true relationships or amplify false signals [@mcelreath2020].

## Ecology

Through modelling the underlying process of spawning arrival and exit, our model produces ecologically informative parameters that provide new insight into the timing of spawning dynamics. These parameters are not available through TAUC or GAUC although some are estimated in previous arrival-exit process models [@szerlong2008; @hilborn1999; @su2001]. In contrast with models that assume a fixed arrival-exit relationship [e.g., @su2001], our model jointly estimates timing parameters using a multivariate normal distribution, enabling flexible, data-informed correlations between arrival and exit. 

Applying this model in the Okanagan sockeye population, we observe considerable interannual variation in some aspects of spawning timing. Arrival timing was surprisingly consistent, ranging by only `r round(max(apply(post$arrival, 2, median))+day_stand - 1) - round(min(apply(post$arrival, 2, median))+day_stand - 1)` days. In contrast, arrival spread varied considerably from year-to-year, ranging from `r round(min(apply(post$arrival_spread, 2, median)), 1)` days in
 `r unique(index_sk$year)[which(apply(post$arrival_spread, 2, median) == min(apply(post$arrival_spread, 2, median)))]` to
 `r round(max(apply(post$arrival_spread, 2, median)),1)` days in
 `r unique(index_sk$year)[which(apply(post$arrival_spread, 2, median) == max(apply(post$arrival_spread, 2, median)))]`.
Exit spread was less variable than arrival spread but showed an overall increasing trend, averaging `r round(mean(apply(post$exit_spread[,1:5], 2, median)),1)` days in the first five years and `r round(mean(apply(post$exit_spread[,(length(unique(index_sk$year))-5):length(unique(index_sk$year))], 2, median)),1)` days in the last five years. We find that in years when spawners arrive later, arrival is spread over a longer period of time, suggesting that the date when the first fish arrive may be relatively constant across years, even if the overall distribution is variable. In Okanagan sockeye, we estimate that exit spread tends to be smaller than arrival spread, which implies that spawner residence time is a function of order of entry (early arrivals stay longer). This is consistent with the assumption in @su2001 and @adkison2001, but here we estimate it directly from the data. This pattern has been identified in other studies, where it has been attributed to the fact that salmon of different age and sex may arrive at different times [@perrin1990; sockeye: @killick1955; pink: @fukushima1997b; @dangel1988].

In addition to improving estimation of the parameters themselves, the multivariate normal correlation structure reveals how arrival and exit processes co-vary. For Okanagan sockeye, we estimate that arrival and arrival spread are positively correlated, indicating that when arrival is more spread out, this pushes the average arrival date later. The other parameters are more weakly associated, although there is some evidence that greater arrival spread (and thus later arrival) is associated with a narrow exit timing, which is consistent with spawners being constrained by biological limits late in the season. 

Extending the analysis of Okanagan sockeye to estimate how the spawning
timing depends on environmental conditions and the spawner's age and sex
composition is an obvious next step. @perrin1990 identified a suite of
potential covariates for residence: spawner density, sex, size, sex
ratio, mean arrival timing (years that are late may have shorter
residence), order of arrival, stream flow, and water temperature. In
Atlantic salmon, older and larger spawners arrive earlier than younger
and smaller spawners, and female salmon arrive earlier in the season
than males [@miettinen2024]. In sockeye, however, males tend to arrive
earlier, and remain alive longer than females [@mathisen1962]. Because
of this, we might expect that years with high proportion fish that
return after one marine winter (age x.1) would have a skewed sex ratio
due to the high prevalence of males (jacks) in that age class – but note
that this population has an unusually high incidence of age x.1 females
(jills) [@major1962].

The four years with the highest proportion of age x.1 fish [2003, 2007,
2019, 2021; @bailey2025] are also years with extremely low spawner
abundances, but do not show anomalous spawning timing @fig-timings. This
is a qualitative comparison; future work could use our model to quantify
how biometric and environmental factors influence each of the timing
parameters of our model, using a casual inference framework
[@cinelli2022].

Notably, the widest spread in arrival times occurred in 2010, which may
be related to the Testalinden Creek dam failure that occurred the
previous spring (June 13, 2010). When this occurred, silt and toxins
were discharged into the Okanagan River [@morhart2010; @tannant2013].
The discharge was downstream of the spawning region but affected Osoyoos
Lake and the Okanagan River below it. Suboptimal conditions there may
have contributed to the increased spread in the timing of spawner
arrivals that year. Of all the model parameters, only arrival spread
appears to be associated with the Testalinden dam failure; how or if the
event directly affected spawners remains unclear.

## Observer accuracy

One persistent challenge with spawner enumeration data is the fact that observations are made with error. In many cases, TAUC methods do not account for observation error at all, although methods exist to do this [@parken2003]. Observer efficiency is typically expected to cause a negative bias; that is, observers only see a fraction of the fish. However, overestimation is also possible. A range of factors, including observer identity, survey method, turbidity, cloud cover, and even spawner abundance potentially affect accuracy of counts. A logical extension of our model would be to allow the overdispersion parameter of the negative binomial distribution in our model, $\phi$, which captures the variability of the counts around around their expected value, to vary as a function of one or more of these factors.

## Further extensions to the model

### Applicability

The spawning behaviour of Osoyoos Lake Sockeye, and the density and
consistency of the visual surveys for those spawners, will not be
typical of other spawner situations. As described below, considering
additional factors (spawner biometrics, short and long-term variability
in spawning habitat) will require a more complicated model than
symmetric Gaussian arrivals and exits. For example, @adkison2001
addressed a multimodal pattern of arrivals. Applying this approach
should be preceded by simulations to determine how bias and precision of
results might deteriorate as survey quality and number decrease,
following our example. From that, the minimal and optimal number
(density and range) of surveys can be compared to extant data. As
explored ( @fig-modcompare ), using fewer annual parameters might lead
to less informative but perhaps more robust results. Application of this
approach for visual surveys in combination with models to analyze
accompanying weir counts and/or tag recoveries will be a valuable
advance.

### Spatial

The model can be extended to represent the counts in individual segments of a spawning region or to include multiple rivers. This will require expanding the hierarchical structure of the model to accommodate variation in parameters across space. Modelling individual segments of a river will be advantageous if spawning timing varies among segments, because aggregating observations at the total-river scale may obscure patterns of arrival and exit. To avoid needing to specify segment-specific priors for abundance, the abundance of each segment could be modelled as a proportion (i.e., a simplex) of the total abundance. Total abundance would still require a prior, as in our current model, and would be estimated directly as the sum across all segments.

### Dead spawners

If available, counts of newly dead spawners could provide further
information to estimate residence. Observations of dead spawners could
be fit with a second likelihood in our model, and linked to the timing
parameters for exits. This would require accounting for the fact that
newly dead spawners accumulate between surveys, so the counts depend on
survey intervals and whether or not dead spawners are marked to ensure
they are only counted once (by tags, by cutting off tails, or
*dead-pitch* removal from the river bank). Some dead spawners will be
consumed by predators and scavengers or drift away, thus not available
to be counted, necessitating a factor to estimate fraction of dead
observed, possibly involving a function of river flow.

### Autocorrelation within and among years

Survey counts within a year are not independent, so it may be appropriate to add an autocorrelation structure to the residuals. We explored this in our model, but found that it did not improve predictive performance when applied to the Okanagan dataset. Furthermore, convergence was poor because autocorrelation and overdispersion in the count data were not well distinguished. We therefore elected to use the simpler model that implicitly assumes independence of consecutive survey counts.

Salmon life cycles result in strong autocorrelations in spawner
abundance. Including this in our model will, conceivably, improve
individual abundance estimates. For sockeye, the four-year cycle
identifiable in @fig-runs is frequently evident [@burgner1991a;
@ricker1950; @ricker1997]: adult returns are predominately the age 1.2
offspring from spawners four years earlier. Because sockeye adults
return two to six years after spawning, not just four, the
autocorrelations are complicated. Extending this model to complete life
histories will require a series of additional estimates (models):
spawners to smolts, smolts to adult returns (SAR), fishing mortality,
and prespawning mortality. These estimates are typically calculated so
this extension is foreseeable.

# Conclusions

This modelling framework estimates the dynamics underlying the spawning
period for salmon, providing annual estimates of key timing parameters
that offer insight into the system's ecology. It also generates annual
estimates of total spawner abundance, with associated uncertainty---the
primary objective of spawner survey programs. This represents a
substantial advancement over conventional TAUC approaches, while
extending the GAUC and arrival-exit process modelling approaches. In
doing so, we provide a way to extract greater insight into the abundance
and ecology of salmon from data that are already routinely collected.
This approach can support the sustainable management of salmon fisheries
by providing more robust abundance estimates and a better understanding
of the factors that govern spawning dynamics. We look forward to results
from challenging this model with differing situations of surveys,
habitats, and species, and to the continued evolution of this approach.

# Acknowledgements

Diligence, care, and persistence by survey field teams from the Okanagan
Nations Alliance (ONA, [syilx.org](https://syilx.org/)) resulted in a
spawners dataset based on unusually thorough and consistent methods;
their commitment to *cause to come back* is inspiring. This paper
commemorates accomplishments of the late Dr. Kim Hyatt who designed,
implemented, and supported this survey program. We thank Karilyn Alex
(ONA) for access and permission regarding this proprietary data,
possible through the data sharing agreement of the Canadian Okanagan
Basin Technical Working Group (COBTWG). We thank Howard Stiff and Athena
Ogden (Pacific Biological Station, Fisheries and Oceans Canada) and
Elizabeth Ng (Four Peaks Environmental Ltd.) for the work required to
reconcile and summarize 25 years of field notes and spreadsheets into
the quality controlled and standardized dataset we received. Dr. James
R. Irvine (DFO Pacific, retired) improved a preliminary version based on
his extensive knowledge of salmon survey methods and related analyses.
We used ChatGPT (model 4o) to proofread specific sections (e.g., mathematical notation and description of the model) of the manuscript that were provided as text excerpts, focusing on grammar,
clarity, and accuracy. All AI generated suggested edits were assessed
and verified for accuracy before being incorporated.

## Competing Interests

The authors declare no competing interests.

## Data Availability

The data used in this study are the property of the Okanagan Nations Alliance. Please contact them directly concerning availability and acknowledgements: [Sylix.org](sylix.org).


# Supplementary Materials {.appendix}

```{=latex}
\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}
```

```{r fig-obsfit, echo = FALSE, warning=FALSE, message=FALSE, fig.align='center', fig.height=5, fig.width=8, fig.cap = "Observed survey counts versus predicted on linear (A) and log (B) scales. Points and error bars show the posterior median and the 66% and 95% credible intervals.",}
spread_draws(fit, live_mu[obs]) %>% 
  left_join(index_sk %>% mutate(obs = 1:n())) %>% 
  ggplot(aes(x = live, y = live_mu, group = obs))+
  geom_abline(slope = 1, intercept = 0)+
  stat_pointinterval()+
  scale_y_continuous(labels = label_number(scale = 1e-3, suffix = "k"), name = "Expected count (thousands)")+
  scale_x_continuous(labels = label_number(scale = 1e-3, suffix = "k"), name = "Observed count (thousands)")+
  theme_bw()+
  
  spread_draws(fit, live_mu[obs]) %>% 
  left_join(index_sk %>% mutate(obs = 1:n())) %>% 
  ggplot(aes(x = live, y = live_mu, group = obs))+
  geom_abline(slope = 1, intercept = 0)+
  stat_pointinterval()+
  scale_y_log10(name = "Expected count")+
  scale_x_log10(name = "Observed count")+
  theme_bw()+
  
  plot_annotation(tag_levels = "A")
```


All scripts and stan model code are available for public access on [GitHub](https://github.com/SiRE-P/Multilevel-Spawners) and [Zenodo](https://support.zenodo.org/help/en-gb): 

# References {.unnumbered}
