---
title: "plt_sim_v2"
author: "JLT, SAA"
date: today
format: pdf
editor: visual
execute:
   output: hold
---

## Adapted from plt_sim.R 2025-05-29

```{r}
#| label: setup
#| echo: false
#| warning: false
library(tidybayes); library(tidyverse); library(scales); library(rstan); 
library(magrittr); library(patchwork);  library(knitr); 
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE) 
# set.seed(123) 
```

## Simulation

### Simulation parameters  
Assign distributions to parameters for spawner behaviour. Also years and days for surveys.

```{r}
#| label: parameters
log_run_mu               <-  10  # lognormal
log_run_sigma            <-   1
arrival_mu               <-  35  # normal  
arrival_sigma            <-   5
arrival_spread_mu        <-   5  # lognormal
arrival_spread_sigma_log <- 0.3
exit_spread_mu           <-   5  # lognormal
exit_spread_sigma_log    <- 0.3
exit_lag                 <-  10  # lognormal
exit_lag_sigma_log       <- 0.3
years                    <-  10
days_min                 <-   1
days_max                 <- 100
rnbinom_size             <-  10  # sampling error
```

### Spawner residence
From cumulative arrived minus cumulative exited for each day.
```{r}
#| label: simulate
    # run size, spawner abund. both params logs
N_total <- round(rlnorm(years, log_run_mu, log_run_sigma)) 
    # 95% 3.1k to 156k
arrival <- rnorm(years, arrival_mu, arrival_sigma)
arrival_spread <- rlnorm(years, log(arrival_spread_mu), arrival_spread_sigma_log)
    # e^.3 = 1.35,  15:50
exit_spread <- rlnorm(years, log(exit_spread_mu), exit_spread_sigma_log)
exit_lag <- rlnorm(years, log(exit_lag), exit_lag_sigma_log)
    # range 5 to 15
output <- sapply(1:years, function(i){
  x <- days_min:days_max
  entered <- pnorm(x, arrival[i], arrival_spread[i])
  exited <- pnorm(x, arrival[i] + exit_lag[i], exit_spread[i])
  present <- entered * (1-exited) * N_total[i]  # only present saved
})

par(tcl=0.2, mgp=c(1,0,0), pty='s',xaxs='i',yaxs='i')
matplot(output, type = 'l', xlab='Day', ylab='Spawner Presence')
axis(3,labels=F);axis(4,labels=F)
```

### Surveys  
Define sampling as a fraction randomly selected of all days. Add negative binomial sampling error. Remove simulated values of zero and sampled observations of zero. 

```{r}
#| label: surveyed
full_surveyed.df <- data.frame(
    N = round(c(output)),
    sampled = rnbinom(n = length(output), mu = output, size = rnbinom_size),
    year = rep(1:years, each = 100),
    day = 1:100)

sampled.df <- full_surveyed.df %>%
  group_by(year) %>%
  sample_frac(size = 0.9)  #  see slice_sample
     # delete true = 0
sampled.df2 <- sampled.df %>% filter( (sampled > 0) & (N > 0) )

p1 <- ggplot(sampled.df2, aes(N, sampled)) + theme_bw()+ 
        geom_point() +
        geom_abline(slope=1, intercept=0) +
        labs(x="Simulated",y="Observed")
p2 <- ggplot(sampled.df2, aes(log10(N), log10(sampled)) ) + theme_bw()+
        geom_point()+
        geom_abline(slope=1, intercept=0) +
        labs(x="log10 Simulated",y="log10 Observed")
p1+p2  # patchwork for plots layout
#ggsave('sampled_df.pdf',width=7, height=7)

ggplot(sampled.df2, aes(x = day, y = sampled)) + theme_bw() +
        geom_point(size=0.5) +       
       labs(x="Day",y="Observed") +
        facet_wrap(~year, scales='free_y')

```

## Fit

### Priors for Stan  
Affects sampling of the parameters space. Four parameters are treated as groups (multilevel) of annual values: *arrival_timing, arrival_spread, exit_lag, exit_spread*. The priors define the distributions for parameters defining distribution of annual values. Annual values are sampled (fitted) from the group distribution, and group parameters are simultaneously fitted, thereby "sharing information" to improve the fit for each annual value. Note *log_runs* is not a group, the same prior applies to each year.

In what follows, *exit_lag_sigma* uses the value for *spread_sigma.* The Stan code renames *count_dispersion* as *log(live_phi)* where *live_phi* is the size parameter for the negative binomial distribution.


```{r}
#| label: tbl-priors
#| tbl-cap: Priors for Stan sampling of parameters space.
priors <- data.frame(prior = c("log_runs_mu", "arrival_mu", "arrival_sigma", "spread_mu","spread_sigma","exit_lag_mu", "count_dispersion"),
                     v1 = c(10, 40,  0, 5, 0, 10,   1),
                     v2 = c( 1,  5, 10, 1, 5,  2, 0.2))
pr_tbl = cbind(priors,
               pdd=c('lnorm','norm','exp','norm','exp','norm','norm'))
kable(pr_tbl,digits=3)

# previously
# v1 = c(10, 40,  0,log(8-1), 0, log(11-1),   1),
# v2 = c( 1,  5, 10,     0.2, 5,       0.3, 0.2))
```

### Data list for Stan  
The priors are part of the Stan data list. Convergence of the HMC sampling chains is improved by providing initial values for *arrival_mu*.
.
```{r}
#| label: tbl-stanDat
sp_dat = list(n_priors = nrow(priors),
              priors = data.matrix(priors[,-1]),
              n_years = max(sampled.df2$year),
              year = as.numeric(factor(sampled.df2$year)),
              day = sampled.df2$day,
              n_obs = nrow(sampled.df2),
              live_counts = sampled.df2$sampled,
              live_phi = 10,
              arrival_mu = arrival_mu)

```

### Stan sampling

```{r}
#| label: stan
mod <- stan_model("spawners_m8.stan")
z <- Sys.time()
fit <- sampling(mod, data = sp_dat,
                chains = 4, cores = 4,
                init = replicate(4, list(arrival_mu = 35), simplify = FALSE))
print(paste0('elapsed time:', format(Sys.time() - z) ))

traceplot(fit)
post <- rstan::extract(fit) # several versions of extract()
```

### Fitted runs.
Spawner abundance estimates presented logged and arith***met***ic plots. Ratio log_10 / log_e = 0.4342945

```{r}
#| label: fittedRuns

run_fit <- spread_draws(fit, log_run[year]) %>%
  filter(.chain %in% c(3:4))
p1 <- ggplot(run_fit, aes(x = year, y = exp(log_run))) + theme_bw() +
      stat_pointinterval()+ labs(x="Year", y= "Run") +
      geom_point(data = data.frame(
        year = 1:years, log_run = log(N_total) ),
        color = 2)
p2 <- ggplot(run_fit, aes(x = year, y = 0.4343*log_run))+ theme_bw()+
      stat_pointinterval()+  labs(x="Year", y= "log_10 Run") +
      geom_point(data = data.frame(
        year = 1:years, log_run = log10(N_total)), 
        aes(x=year, y=log_run), color = 2)
p1+p2
```

### Fitted parameters  
The annual values for spawner *arrival_timing, arrival_spread, exit_lag*, and *exit_spread* are presented as marginal posterior distributions (confidence limits) and compared to the simulated value. The goal is finding any consistent bias.

```{r}
#| label: fittedParams

p1 <- spread_draws(fit, arrival[year]) %>%
  filter(.chain %in% c(3:4)) %>%
  ggplot(aes(x = year, y = arrival)) + theme_bw() +
  stat_pointinterval()+ labs(x="Year", y="Arrival Timing") +
  geom_point(data = data.frame(year = 1:years, arrival = arrival), color = 2)

p2 <- spread_draws(fit, arrival_spread[year]) %>%
  filter(.chain %in% c(3:4)) %>%
  ggplot(aes(x = year, y = arrival_spread)) + theme_bw() +
  stat_pointinterval()+ labs(x="Year", y="Arrival Spread") +
  geom_point(data = data.frame(year = 1:years, arrival_spread = arrival_spread), color = 2)

p3 <- spread_draws(fit, exit_lag[year]) %>%
  filter(.chain %in% c(3:4)) %>%
  ggplot(aes(x = year, y = exit_lag)) + theme_bw() +
  stat_pointinterval()+ labs(x="Year", y="Exit Lag") +
  geom_point(data = data.frame(year = 1:years, exit_lag = exit_lag), color = 2)

p4 <- spread_draws(fit, exit_spread[year]) %>%
  filter(.chain %in% c(3:4)) %>%
  ggplot(aes(x = year, y = exit_spread)) + theme_bw() +
  stat_pointinterval()+ labs(x="Year", y="Exit Spread") +
  geom_point(data = data.frame(year = 1:years, exit_spread = exit_spread), color = 2)

p1+p2+p3+p4
```

## Spawner presence
The pattern of spawner distribution is recreated from the fitted parameters and presented with 95% confidence limits. This is compared to the observations.

```{r}
#| label: spawnCurves

spawn_curves.df <- data.frame()
for(j in 1:ncol(post$log_run)){
  live <- sapply(1:100, FUN = function(x) {
    entered <- pnorm(x, post$arrival[,j], post$arrival_spread[,j])
    exited  <- pnorm(x, post$arrival[,j] + post$exit_lag[j],
                     post$exit_spread[,j])
    exp(post$log_run[,j] + log(entered - (entered * exited)))
  })

  spawn_curves.df <- bind_rows(spawn_curves.df,
    data.frame(year = j,
      day  = 1:100,
      fish = apply(live, 2, median),
      l95  = apply(live, 2, quantile, probs = 0.025),
      u95  = apply(live, 2, quantile, probs = 0.975)))
}
```

```{r}
#| label: fig-spawnCurves
#| fig-cap: Estimated spawner presence, with 95% CL, from fitted parameters. Observed values are points.

ggplot(spawn_curves.df, aes(x = day, y = fish))  + theme_bw() +
  geom_line()+
  geom_ribbon(aes(ymin = l95, ymax = u95), color = NA, alpha = 0.2)+
  facet_wrap(~year, scales = "free_y") +
  geom_point(data = sampled.df2, aes(x = day, y = sampled), size = 0.8) +
  scale_y_continuous(
    labels = label_number(scale = 1e-3, suffix = "k"),
    name = "Spawners present (thousands)", expand = expansion(mult = c(0, 0.02))) +
  theme(strip.background = element_rect(color="NA", fill="NA"))+
  scale_x_continuous(
    labels = ~ format(as.Date(.x, origin = "2023-12-31"), "%b"),
    breaks = as.numeric(as.Date(c("2024-09-01", "2024-10-01", "2024-11-01")) - as.Date("2023-12-31")), name = "") +
  theme(panel.grid.minor = element_blank())

ggsave('spawn_curves.pdf',width=14, height=14)

```
