---
title: "plt_sim version 4"
author: "JLT, SAA"
date: today
format: pdf
execute:
   output: hold
   echo: false
---

## Goal  

Tables and plots showing fits for runs and the four spawning parameters as a function of:  
(1) changing exit_lag, to prove that is reliably recovered, and  
(2) sampling intervals to investigate results for sub-optimal surveys.  
Uses model **spawner_m8.stan** to fit model to simulated data plus sampling error. Results are  
(1) median of the ratio fitted/actual, to describe consistent % bias, and  
(2) median absolute difference, MAD, of that ratio to estimate % precision

## History  

-   From plt_sim version 3.3 2025-07-09.
-   From plt_sim version 3.2.JLT 2025-06-28.
    -   Bug fix, GitHub.
-   From plt_sim version 3.1 2025-06-23.
    -   Uses spawners_m8_beta.stan
-   From plt_sim_v3.qmd 2025-06-23.
    -   This and prev use spawners_m8.stan
-   From plt_sim_v2.qmd 2025-06-14
-   From plt_sim.R 2025-05-29
-   From Simulation-for-Supplementary 2025-05-20

```{r}
#| label: setup
#| include: false
#| echo: false
#| warning: false
library(tidybayes); library(tidyverse); library(scales); library(rstan); 
library(magrittr); library(patchwork);  library(knitr); 
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE) 
set.seed(104) # reproduce random samples.
```

## Difference from preceding  
Survey dates are evenly distributed sequences, not random.\
Scenarios repeated and fitted with differing intervals between surveys: 1, 3, 6, 9 days.\
Scenarios repeated and fitted with differing fixed values for exit_lag: 5,10, 15 days.\
20 years instead of 12.\
TAUC included, with analysis of bias and precision\
updates to fig- and tbl- captions.\
.pdf output for inclusion in primary.\

# Simulation

## Distributions for simulation parameters  
Assign distributions to parameters for runs and spawner behaviour. Set years and days for surveys. Parameters as per JLT code: spawners_m8.stan

```{r}
#| label: parameters
#| echo: true
# revised to match JLT email 2025-07-28, 13:39.
log_run_mu               <-  10    # log value for lognormal PDD, 
log_run_sigma            <-   1    # log value for lognormal PDD, 
arrival_timing_mu        <-  35    # normal  
arrival_timing_sigma     <-   1    # 95% days 33 to 37 
arrival_spread_mu        <-   4    # normal
arrival_spread_sigma     <-   0.2  # 95% 3.6 to 4.4
exit_lag_mu              <-  10    # normal 
exit_lag_sigma           <-   0.5  # 95% 9 to 11
exit_spread_mu           <-   3    # normal. should be < arrival_spread_mu
exit_spread_sigma        <-   0.2  # 95%  2.6 to 3.4
years_n                  <-  20    # number of years, 4x3 plot
days_n                   <- 100    # initial number of days, checked below.
samp_accuracy            <-   0.15 # rnorm(n, 1, 0.15) 
    # 15% sampling error, +/- 95% = 0.3 is a lot. Likely under not over. 
gaps                     <- c(1,3,5,7) # days between surveys. survey_gaps identifies scenarios.
    # live_phi is the size parameter for the negative binomial distribution
live_phi                 <-  10    # sampling error (version 3.1)
    # derived
days  = 1:days_n                   # overrides lubridate::days()
years = 1:years_n                  # overrides lubridate::years()
sim = c(arrival_timing_mu, arrival_timing_sigma,
        arrival_spread_mu, arrival_spread_sigma,
        exit_spread_mu, exit_spread_sigma, 
        exit_lag_mu, exit_lag_sigma)
saveRDS(sim, 'sim.rds') # save for use after rendering
```

## Draws for annual values  
Random samples from preceding distributions to set annual values for run size and for the four spawning parameters. The likely ranges for draws (99% CL) are in @tbl-CL99. The resulting sets of `r years_n` draws are in @tbl-draws. Run sizes are lognormal: occasional unusually large runs. The 99% CLs for the parameter distributions are in @tbl-CL99, and the annual draws are in @tbl-draws. The draws are checked @tbl-checkSpreads for a possible pathology where exits are complete before arrivals are complete. Finally, the annual distributions of spawner presence are displayed @fig-sampled as well as log presence @fig-logSampled.

```{r}
#| label: tbl-CL99
#| tbl-cap: Range (99%) of random draws for annual values of run size and spawning parameters.
p99 = c(0.01,0.99) # +/- 2.326348 stdevs
CL99= data.frame(
  Parameter = c('run','arrival_timing','arrival_spread','exit_lag','exit_spread'),
  p01 = c(qlnorm(p99[1],log_run_mu, log_run_sigma) %>% round,
          qnorm(p99[1],arrival_timing_mu, arrival_timing_sigma) %>% round(2),
          qnorm(p99[1],arrival_spread_mu,arrival_spread_sigma) %>% round(2),
          qnorm(p99[1],exit_lag_mu, exit_lag_sigma) %>% round(2),
          qnorm(p99[1],exit_spread_mu, exit_spread_sigma) %>% round(2)),
  p99 = c(qlnorm(p99[2],log_run_mu, log_run_sigma) %>% round,
          qnorm(p99[2],arrival_timing_mu, arrival_timing_sigma) %>% round(2),
          qnorm(p99[2],arrival_spread_mu,arrival_spread_sigma) %>% round(2),
          qnorm(p99[2],exit_lag_mu, exit_lag_sigma) %>% round(2),
          qnorm(p99[2],exit_spread_mu, exit_spread_sigma) %>% round(2))
)
kable(CL99, align='r')
```

```{r}
#| label: tbl-draws
#| tbl-cap: Random draws for annual values of run size and spawning parameters.
draws = data.frame(  year = years,
run             = rlnorm(years_n,  log_run_mu, 
                    log_run_sigma) %>% round,
arrival_timing  = rnorm(years_n, arrival_timing_mu,
                    arrival_timing_sigma) %>% round(2),
arrival_spread  = rnorm(years_n,arrival_spread_mu,
                    arrival_spread_sigma)  %>% round(2),
exit_lag        = rnorm(years_n, exit_lag_mu, 
                    exit_lag_sigma) %>% round(2),
exit_spread     = rnorm(years_n,exit_spread_mu, 
                     exit_spread_sigma) %>% round(2) )

kable(draws, digits=1, align='c')
saveRDS(draws, 'draws.rds') # save for use after rendering
```

```{r}
#| label: tbl-checkSpreads
#| tbl-cap: Typically arrival_spread is greater than exit_spread. If less than, and with small exit_lag, this can lead to late spawners exiting before arriving. This is checked as lag between the day 99% have arrived to the day 99% have exited.
    # Days after 99% arrived until 99% exited?
test_spread = data.frame(
  Year = years,
  Spread_Difference = draws$arrival_spread - draws$exit_spread,
  In_Before_Out = 
      (qnorm(0.99,draws$exit_lag,draws$exit_spread) -
       qnorm(0.99,       0,draws$arrival_spread)) )
kable(test_spread, digits=1, align = 'r')
```

```{r}
#| label: present
# apply annual draws for to simulate spawners present.
present <- sapply(1:years_n, function(i){
  entered <- pnorm(days, draws$arrival_timing[i],  draws$arrival_spread[i])
  exited  <- pnorm(days, draws$arrival_timing[i] + draws$exit_lag[i], draws$exit_spread[i])
  present <- (entered * (1-exited) * draws$run[i]) # zero if out before in
  }) 
dat_sim = data.frame(
  year    = rep(years,each=days_n),
  day     = rep(days,years_n),
  present = present%>% as.vector %>% round)
    # first and last days of presence over all year.
a= dat_sim$day[dat_sim$present > 0]
day_first = min(a)
day_last  = max(a) 
cat('range of days with spawners present:', day_first, "to", day_last, '\n')
rm(a)
```

```{r}
#| label: fig-present
#| fig-cap: Time series of spawners present in each year.
#| warning: false

ggplot(dat_sim, aes(day, 1e-3*present, color=factor(year) )) +
  theme_bw() + theme(panel.grid.minor = element_blank()) +
  geom_line() +
  scale_x_continuous(limits=c(day_first, day_last)) +
  labs(x="Day", y="Spawners Present ('000)" ) +
  theme(legend.position = "none") 
ggsave('spawners present.pdf',width=7, height=7)
```

## Peak live   
As a relative index of spawner abundance, how does peak live prsence compare to the true run size? Regression through the origin @fig-peakLive provides an $R^2$ value, and the fitted slope is compared to slope=1.

```{r}
#| label: fig-peakLive
#| fig-cap: Peak live spawners in each year compared to true run size
pltdat= data.frame(
    PL= by(dat_sim$present/1000, dat_sim$year, max) %>% as.vector,
    run=draws$run / 1000) 
a = lm(PL ~ 0 + run, data=pltdat)
a_pred = predict(a)
r2 = 1 -  sum((a_pred - pltdat$PL)^2) / sum((a_pred - mean(a_pred))^2)  
slope= summary(a)$coefficients[1,1] %>% round(2) # slope of regression line

ggplot(pltdat, aes(run, PL)) +
  theme_bw() + theme(panel.grid.minor = element_blank()) +
  geom_point() +
  stat_smooth(method="lm", formula=y ~ 0+x ) +
  geom_abline(slope=1, intercept=0, color='red') +
  labs(x="Run ('000)", y="Peak Live ('000)") +
  annotate("label", x=0.5*max(pltdat$run), y=max(pltdat$PL), 
    label=paste("slope = ",slope, ",  R^2 = ", round(100*r2),"%", sep=''), size=5)

ggsave('peak_live_vs_run.pdf',width=7, height=7)
```


## Scenarios  
A scenario has surveys at fixed _survey_gaps_, preset as _gaps_ (typically 1, 3, 5, 7 days). Same dates every year, but note spawner _arrival_timing_ varies by year. Proportional sampling error (normal) is applied to generate observations, @fig-samplError. Surveys removed if zero for simulated or sampled observations.

```{r}
#| label: surveyed

# apply proportional sampling error
dat_sim %<>% mutate(sampled = present*rnorm(nrow(dat_sim),1, samp_accuracy))
dat_sim %<>% mutate(sampled = floor(sampled)) # truncate: 0.9 fish -> 0 fish.
# determine survey dates with 1, 3, 5, 7 day gaps. 
# save four survey scenarios. Same dates all years
days_surveyed <- list(
  seq(day_first,day_last, by=2), 
  seq(day_first,day_last, by=4), 
  seq(day_first,day_last, by=6), 
  seq(day_first,day_last, by=8) )
# create long form data frame for scenarios
sampled.df <- data.frame()
for(i in 1:length(days_surveyed)){
  sampled.df <- bind_rows(sampled.df, 
    dat_sim %>% filter(day %in% days_surveyed[[i]]) %>%
    mutate(survey_gaps=gaps[i] ))
}
# delete if present =< 0 or sampled =< 0 
sampled.df %<>% filter((sampled > 0) & (present > 0))
# hist(sampled.df$sampled/sampled.df$present, breaks=13)
# range(sampled.df$sampled/sampled.df$present) # 0.5, 1.46
```

```{r}
#| label: fig-samplError
#| fig-cap: Comparison of spawners observed to spawners present.
#| fig-subcap: Sampling error is proportional to the number of spawners present. Note log(0)=1

plt_dat = filter(sampled.df, survey_gaps == 1 )

p1 <- ggplot(plt_dat, aes(1e-3*present, 1e-3*sampled)) +
  theme_bw() + theme(panel.grid.minor = element_blank()) +
  geom_point(size=0.33) +
  geom_abline(slope=1, intercept=0, color='red') +
  labs(x="Present ('000)",y="Sampled ('000)")
p2 <- ggplot(plt_dat, aes(log10(present), log10(sampled)) ) +
  theme_bw() + theme(panel.grid.minor = element_blank()) +
  geom_point(size=0.33)+
  geom_abline(slope=1, intercept=0, color='red') +
  labs(x="log10 Present",y="log10 Sampled")
p1+p2  # patchwork for plots layout
ggsave('sampled_df.pdf',width=7, height=7)
```

## Observations  

Derived observations (@fig-sampled) show scatter near day of peak abundance (mode). Viewed as logs (@fig-logSampled) to show negative skew when arrival_spread is greater than exit_spread (typical), and positive skew when smaller. 

```{r}
#| label: fig-sampled
#| fig-cap: Annual time series of spawners from surveys with 1 day gaps (black dots) and 4 day gaps (bluecircles).
ggplot(plt_dat, aes(x = day, y = 1e-3*sampled)) + theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  geom_point(size=0.5) +
  geom_point(data=filter(sampled.df, survey_gaps == 7), 
    aes(x=day, y=1e-3*sampled), shape=1, color='blue') +
  labs(x="Day",y="Sampled ('000)") +
  facet_wrap(~year, scales='free_y') +
  theme(panel.grid.minor = element_blank()) +
  theme(strip.background = element_rect(color="NA", fill="NA") )
ggsave('sampled.pdf',width=7, height=7)
```

```{r}
#| label: fig-logSampled
#| fig-cap: Annual time series of logged observations.
#| fig-subcap: Reveals skewed distributions.
ggplot(plt_dat, aes(x = day, y = log10(sampled))) + theme_bw() +
  geom_point(size=0.5) +
  geom_point(data=filter(sampled.df, survey_gaps == 7), 
    aes(x=day, y=log10(sampled)), shape=1, color='blue') +
  labs(x="Day",y="Sampled (log_10)") +
  facet_wrap(~year, scales='free_y') +
  theme(panel.grid.minor = element_blank()) +
  theme(strip.background = element_rect(color="NA", fill="NA") )

```




# Fit

## Priors for Stan  
Bayesian priors (@tbl-priors) affect MCMC (Markov Chain Monte Carlo, or related) samples of the parameters hyperspace. The spawning parameters: *arrival_timing, arrival_spread, exit_lag, exit_spread* are treated as multilevel, groups of annual values.  

In what follows, the prior *exit_lag_sigma* uses the value for *spread_sigma.* The Stan code renames *count_dispersion* as *log(live_phi)* where *live_phi* is the size parameter for the negative binomial distribution.  

```{r}
#| label: tbl-priors
#| tbl-cap: Priors for Stan sampling of parameters space.
priors <- data.frame(
  prior = c("log_runs_mu", "arrival_mu", "arrival_sigma", "spread_mu","spread_sigma","exit_lag_mu", "count_dispersion"),
  pdd=c('lnorm','norm','gamma','lnorm', 'gamma','lnorm','exp'),
  v1 = c(10,     40,      0,    log(8-1),  0,    log(10),   1),
  v2 = c( 1,      2,      1,      0.2,    10,     0.3,     0.2) )
kable(priors,digits=2)

```

```{r}
#| label: taucJLT
#| eval: false 
    # determine TAUC, Trapezoidal Area Under Curve, for each year
    # author: JLT 
TAUC = function (){
    # df is a data frame with columns: year, day, sampled
  df %<>% group_by(year) %>% 
    summarise(TAUC = sum(0.5 * (lead(sampled,1) + sampled) * (lead(day,1) - day)) )
  return(df)
}
```

## Data list for Stan  
The priors are part of the Stan data list. Convergence of the HMC sampling chains is improved by providing initial values for *arrival_mu*.\
This data list is modified in the loop for the four scenarios of survey density.

```{r}
#| label: stanDat
#| echo: true
sp_dat_0 = list(
  n_priors     = nrow(priors),
  priors      = data.matrix(priors[,3:4]),
  n_years     = years_n, 
  arrival_mu  = arrival_timing_mu )

#  n_obs       = nrow(sampled.df),
#  year        = sampled.df$year, 
#  day         = sampled.df$day,
#  live_counts = as.integer(sampled.df2$sampled),
  
```

## Stan compiled  

```{r}
#| label: stanCompile
    # file.exists("../stan/spawners_m8.stan") 
mod <- stan_model("../stan/spawners_m8.stan")
```

## Fits by scenario  

```{r}
#| label: fits
z <- Sys.time() 
    # create data.frames to hold results of multiple fits.
group_medians= data.frame()  # fitted medians group parameters
annual_medians= data.frame() # fitted medians annual parameters
    # loop through the four scenarios of survey gaps
for (j in gaps){   # 1,3,5,7 
    # insert scenario data into stan data  
  k = sampled.df$survey_gaps == j  # which rows
  sp_dat = list_modify(sp_dat_0,
      n_obs = sum(k),
      year  = sampled.df$year[k],
      day   = sampled.df$day[k],
      live_counts= sampled.df$sampled[k] %>% as.integer )

  fit <- sampling(mod, data = sp_dat,
    chains = 4, cores = 4, iter=1500, warmup=1000,
    init = replicate(4, list(arrival_mu=runif(1,32,37)), simplify=FALSE),
    open_progress=TRUE, control=list(max_treedepth=20) ) 
 
  traceplot(fit, nrow=5, ncol = 2 ) +  # defaults to first 10 parameters
    theme(legend.position = "none") 
    ggsave(paste0('traceplot_scenario_', j, '.pdf'))
  
      # medians of posterior distn (fits). rows: 20 years. accumulate, save: .rds
  meds = summary(fit)$summary[c(1:9, 10:29,110:129, 130:149, 150:169, 170:189), 6] 
  group_medians= rbind(group_medians, data.frame(
        med = meds[1:8]) ) # drop live_phi 
  annual_medians = rbind(annual_medians, data.frame(
        run        = exp(meds[10:29]),  # was log_run
        arrival_timing = meds[30:49],
        arrival_spread = meds[50:69],
        exit_spread    = meds[70:89],
        exit_lag       = meds[90:109]) ) # note order of columns
}

saveRDS(group_medians,  "group_medians.rds") # save for use after rendering
saveRDS(annual_medians, "annual_medians.rds") 

cat(paste0(' *** \n Fits completed after : ', format(Sys.time() - z) ))
```

## Medians  
Fitted medians for group values @tbl-groupMeds and annual values @tbl-annualMeds. From the fit for each scenario. Saved as group_medians.rds

```{r}
#| label: tbl-groupMeds
#| tbl-cap: Fitted medians for group parameters,  by scenario (survey gaps).
    # 4 groups, 2 fitted params each.
# group_medians = readRDS("group_medians.rds") # dataframe:1 col,8 param,4 gap
# sim = readRDS("sim.rds") # 1 col, 8 params
g_names= c('arrival_timing_mu', 'arrival_timing_sigma',
           'arrival_spread_mu', 'arrival_spread_sigma',
           'exit_spread_mu', 'exit_spread_sigma', 
           'exit_lag_mu', 'exit_lag_sigma')
s_names = c('Sim', 'Gaps=1','Gaps=3','Gaps=5','Gaps=7') # true and 4 scenarios
gm  = matrix(unlist(group_medians), nrow=8,ncol=4) %>% 
    as.data.frame %>% `rownames<-`(g_names) 
row.names(gm) = g_names # 8 rows
gm[3:8,] = exp(gm[3:8,]) # exported as log values
gm = cbind(sim %>% as.data.frame, gm) %>% # add simulation values
    `colnames<-`(s_names) 
kable(gm, digits=2)
```

```{r}
#| label: tbl-annualMeds
#| tbl-cap: Fitted medians for annual parameters, by scenario (survey gaps).
    # annual_medians = readRDS("annual_medians.rds") 
    # 4 scenarios, 20 years: 80 cols
p_names = c('run', 'arrival_timing', 'arrival_spread', 'exit_spread', 'exit_lag')
annual_medians = cbind(year = rep(1:years_n, 4),
                       gap= rep(c(1,3,5,7), each = years_n),
                       annual_medians)
    # reorg draws to match annual_medians. years_n = 20 rows, 6 cols
d1 = data.frame(year=1:years_n, gap=rep(0,years_n), draws[, -1]) 
    # draws to be comparable to scenarios: is 20 rows, 7 cols
    # add draws to top of annual_medians, but rename that

annual_meds= rbind(d1, annual_medians) # stack true values above fitted values
j = with(annual_meds, order(year, gap)) # rows: gap within year
annual_meds = annual_meds[j,]

kable(annual_meds, digits=2, row.names=FALSE) 
```

## Bias and Precision
For annual values, compare fit to true as ratio: observed over expected. Apply over all years for each scenario of survey gaps. Accuracy is two-dimensional: bias (median) and precision(MAD, median absolute deviation). The result is @tbl-BP and @plt-BP. 


```{r}
#| label: medmad
    # redone to use annual_meds that includes true values (draws).
    # 7 cols: year gap run arrival_timing arrival_spread exit_spread exit_lag
MedMad = function (meds, years_n=20, gaps_n=5){
    # ratio of fitted/true, by year, by gap, by parameter
    # then medians and MADs of annual ratios, by gap, by parameter
    # step 1. ratios of fit to true
    x = meds # copy to hold interim ratios
    c_names =colnames(meds)[3:7]
    for (k in 3:7) {  # 5 parameters
        for(i in 1:years_n) { 
            # for each gap, for each parameter
            m= (i-1)*gaps_n + (1:gaps_n)
            z = meds[m, k] # fits for one year for one parameter
            x[m, k] = z/z[1] # ratio to first
        }
    } 
    # step 2. delete ratio of true to true, where gap == 0
    x = x[x$gap != 0, ]  
    gaps_n = gaps_n - 1 
    x = x[order(x$gap),3:7 ]  # order by gap. drop year and gap.
    # step 3. median and MAD over all years, by gap, by parameter
    x_med = x_mad = matrix(nrow=gaps_n, ncol=5) %>% as.data.frame  # hold result 
    for (k in 1:5) {  # 5 parameters
        for(i in 1:gaps_n) { 
            # for each gap, for each parameter
            m= (i-1)*years_n + (1:years_n)
            z = x[m, k] # ratios all years for gap i, param k.
            x_med[i, k] = median(z)
            x_mad[i, k] = mad(z)
        }
    } 
    # step 4. combine medians and MADs
    z= cbind(x_med,x_mad) %>% `colnames<-`(c(paste0('med_', c_names), paste0('mad_', c_names)) )
    return(z)
} 

annual_mm = MedMad(annual_meds)
saveRDS(annual_mm, "annual_med_mad.rds") # for use after rendering
```


```{r}
#| label: tbl-BP
# tbl-cap: From annual ratios of fit/draw, bias and precision as percent, by scenarios of survey gaps.
# annual_mm = readRDS("annual_med_mad.rds") # 10 cols, 4 rows = gaps
# p_names = c('run', 'arrival_timing', 'arrival_spread', 'exit_spread', 'exit_lag')
# s_names = c('Sim', 'Gaps=1','Gaps=3','Gaps=5','Gaps=7') # true and 4 scenarios
bias = ((annual_mm[,1:5 ]-1) * 100) %>% round %>%
    `colnames<-`(p_names) %>% t %>% `colnames<-`(s_names[-1])
prec = (annual_mm[,6:10] * 100) %>% round %>% 
    `colnames<-`(p_names) %>% t %>% `colnames<-`(s_names[-1])

kable(bias, caption = 'Bias (percent) from median of fitted annual parameters divided by the initial simulation values, as a function of increasing survey gaps (days).')

kable(prec, caption = 'Precision (percent) from MAD of fitted annual parameters divided by the initial simulation values, as a function of increasing survey gaps (days).')
```


## Plot relative bias and precision  

```{r}
#| label: plt-BP
#| fig-cap: Relative bias and precision of fitted annual parameters divided by the initial simulation values, as a function of increasing survey gaps. Values are  the difference from survey gaps of 1 day.

Rel_to_first = function(a){
    for(j in 1:nrow(a)) a[j,]=a[j,]-a[j,1];return(a)}

plt_dat1 = Rel_to_first(bias) %>%
          rbind(gaps=c(1,3,5,7),.) %>% t %>% as.data.frame %>%
          pivot_longer(cols = 2:6, names_to = "Parameter") 

p1 = ggplot(plt_dat1, aes(x=gaps, y=value, color=Parameter)) +
     theme_bw() + theme(panel.grid.minor = element_blank()) +
     geom_point() +     geom_line() +
     scale_x_continuous(limits = c(0,8), breaks=c(1,3,5,7) ) +
     labs(x="Surveys Gap (days)", y="Relative Bias (%)") +
     theme(legend.position = "none")  # just for LHS plot, not RHS plot

plt_dat2 = Rel_to_first(prec) %>%
           rbind(gaps=c(1,3,5,7),.) %>% t %>% as.data.frame %>%
           pivot_longer(cols = 2:6, names_to = "Parameter") 

p2 = ggplot(plt_dat2, aes(x=gaps, y=-value, color=Parameter)) +
     theme_bw() + theme(panel.grid.minor = element_blank()) +
     geom_point() +     geom_line() +
     scale_x_continuous(limits = c(0,8), breaks=c(1,3,5,7) ) +
     labs(x="Surveys Gap (days)", y="Relative Loss of Precision (%)") 

p1+p2 
ggsave('relative bias and precision.pdf')
```

## TAUC from survey obs  
Algorithm from Millar and Jorday 2013.  
Where $d$ is day and $s_j$ is the sampled value for day $d_j$, the trapezoidal area under the curve and its variance are:
$$
\begin{aligned}
TAUC &= 0.5  \sum_{j=2}^{n-1}  (d_{j+1} - d_{j-1}) * s_j \\
\text{var} (TAUC)  & = 0.25 \sum_{j=2}^{n-1}(d_{j+1} - d_{j-1})^2 * \text{var}(s_j)
\end{aligned}
$$
This requires $ s_{j=1} = 0  \text{ and } s_{j=n}=0$ to be added if not present.
The function *Trapezoid()* calculates TAUC and var(TAUC) for each year, then *TAUC()*    applies Trapezoid() by year by survey gaps (@tbl-TAUC).

```{r}
#| label: tauc
    # determine TAUC, Trapezoidal Area Under Curve, and standard dev of that estimate,
    # for each year. uses sampled.df 
    # author: SAA (notice inelegance cf JLT?)
TAUC = function (x,s){
    # x is a data frame with columns:year, day, sampled, survey_gaps
    # s is stdev is proportional Precision, observed = true * rnorm(n,1,samp_accuracy)
    # so var(sampled) = s^2
    # from Millar and Jorday 2013
  Trapezoid = function(x,s) {
      # trapezoidal area under curve and variance thereof from known sampling error
      # data is extended by leading and trailing days with sampled = 0, 
      # at dates t[0] = t[2] - 2*t[1] and t[n+1] = 2*t[n] - t[n-1] 
    day = x$day; sampled = x$sampled
    sampled = c(0,sampled,0)
    n=length(day)
    day = c( 2*day[1]-day[2], day, 2*day[n] - day[n-1] )
       # (90,100,110,120) -> (80,90,100,110,120,130) 
    n = n+2 # extended   
    y = yv = 0     # trapezoid area and variance
    for (j in 2:(n-1)) {
        dx = 0.5* (day[j+1]-day[j-1])
        y =  y  + dx   * sampled[j]
        yv = yv + dx^2 * (s*sampled[j])^2 # variance from accuracy 
    }
    return( c(y, sqrt(yv)) ) # return area and stdev
  }
    # apply Trapezoid by year and survey_gaps 
  tauc = by(
            data = x[,c('day', 'sampled')], 
            INDICES = list(j1=factor(x$survey_gaps), j2=factor(x$year)),
            Trapezoid, s = s)  %>% 
        unlist %>% matrix(ncol=2, byrow=TRUE) %>% as.data.frame 
  return(tauc)   
}

MedMad_TAUC = function (tauc, run){
    # tauc is a data frame with columns: year, day, sampled, survey_gaps
    # draws is the data frame of annual draws
  MM = function(x, run){
      med_r= median(x/run)  # median of ratio TAUC/run
      mad_r= mad   (x/run)  # MAD of ratio
      return(c(med_r,mad_r) ) # return median and MAD
  }
  mm_t = by (data = tauc[,1],  # run$TAUC
          INDICES = list(pattern=factor(tauc$survey_gaps)),
          MM, run=draws$run) %>% unlist %>% as.vector
    return(mm_t)
} 

```

## TAUC Precision  
Determine *reside*, the effective residence as the ratio of TAUC to true run, using linear regression through origin for the scenario _survey_gaps_ = 1. Then _run_tauc_ = TAUC/reside. Note _reside_ is constant across years. The function *MedMad_TAUC()* calculates the median and MAD of run_tauc/run to estimate relative bias and relative Precision (@tbl-relativeTAUC). Relative means compared to surveys with 1 day gap = two days apart. Bias is relative to 100% (no bias), median ratio 99% is -1% bias. Precision is loss of precision, a larger MAD is less precise.

```{r}
#| label: tbl-TAUC
#| tbl-cap: Bias (median) and precision (MAD) of run size from TAUC compared to true run size before sampling error, by survey gaps.

tauc = TAUC(sampled.df, samp_accuracy) 
colnames(tauc) = c('TAUC', 'TAUC_sd') # rename columns
    # determine residence from TAUC/draws$run
reside = summary(lm(tauc[seq(1,80,4),1] ~ 0 + draws$run[1:20]))$coefficients[1:2]
cat("effective residence = ", reside[1], " days (sd=", reside[2], ")\n", sep='')
a = tauc[seq(1,80,4),1] / draws$run[1:20]
cat("TAUC/run: median = ", round(median(a),2), " (mad = ", round(mad(a),2), ")\n", sep='')
cat("TAUC/run: mean   = ", round(mean(a),2), " (sd=", round(sd(a),2), ")\n", sep='')
cat("TAUC/run: range  = ", range(a)[1] %>% round(2), " to ", range(a)[2]%>% round(2), "\n", sep='')

tauc = tauc / reside[1] # divide by residence to get run size and stdev of run
# tauc is ordered by survey gaps within years. add these.
tauc$year=rep(1:years_n, each=4); 
tauc$survey_gaps=rep(gaps, years_n) # gaps set in Simulation above

mm_t = MedMad_TAUC(tauc, draws$run) %>% `*`(100) %>% round(1) %>%
       matrix(ncol=2, byrow=TRUE) %>% as.data.frame %>% `colnames<-`(c('Median', 'MAD'))
mm_t$Survey_Gaps = c(1,3,5,7) # add survey gaps
kable(mm_t[,c(3,1,2)], digits=2, align='c', col.names = c('Survey Gaps (days)', 'Bias (%)', 'Precision (%)') )

```

```{r}
#| label: tbl-relativeTAUC
#| tbl-cap: "Relative bias and relative loss of precision of TAUC run divided by actual run, compared to survey gaps = 1 day."

mm_t_r  = mm_t  # to hold relative median and MAD for TAUC_run compared to gap=1
for (j in 1:2){ mm_t_r[ ,j] = (mm_t[ ,j] -  mm_t[1,j]) } 
mm_t_r$Survey_Gaps = gaps # add survey gap
mm_t_r$MAD %<>% `*`(-1)  #  to be loss of precision (larger MAD less precise) 

kable(mm_t_r[,c(3,1,2)], digits=2, align='c', col.names = c('Survey Gaps (days)', 'Relative Bias (%)', 'Relative Precision (%)') )
```

```{r}
#| label: fig-relTAUC
#| fig-cap: Relative bias and precision of TAUC run size as a ratio to actual run size. Values are relative to  surveys with 1 day gaps.

pltdat= data.frame(Survey_Gaps = rep(mm_t_r$Survey_Gaps,2),
                   Type = rep(c('Bias (median)','Precision (MAD)'), each = nrow(mm_t_r)),
                   Value = c(mm_t_r$Median, mm_t_r$MAD) )

ggplot(pltdat,  aes(x=Survey_Gaps, y= Value, colour=Type)) +
     theme_bw()+ theme(panel.grid.minor = element_blank()) +
     geom_point() +    geom_line() +
     scale_x_continuous(breaks=c(1,3,5,7) ) +
    labs(x="Surveys Gap (days)", y="Relative Bias and Precision (%)")

 ggsave('TAUC relative loss of Precision.pdf')

```


## TAUC details
@fig-badlySampled is to clarify the pattern from survey results when survey_gaps is large (7 days). @fig-TAUCvsDraw compares run from TAUC to true run, contrasting (?) results from survey_gaps = 1 day and 7 days.

```{r}
#| label: fig-badlySampled
#| fig-cap: Annual time series of spawners from surveys with 7 day gaps
ggplot(data= filter(sampled.df, survey_gaps == 7 ), aes(x = day, y = 1e-3*sampled)) + theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  geom_point() + geom_line() +
  labs(x="Day",y="Sampled ('000)") +
  facet_wrap(~year, scales='free_y') +
  theme(panel.grid.minor = element_blank()) +
  theme(strip.background = element_rect(color="NA", fill="NA") )
``` 

```{r}
#| label: fig-TAUCvsDraw
#| fig-cap: TAUC run size compared actual run size. The red line is slope = 1, intercept = 0. Black points are survey_gaps = 1 day, blue points are survey_gaps = 7 days.
    # match runs to tauc 
plt_dat = data.frame(
    survey_gaps = factor(rep(c(1,7), each=years_n)), 
    draw_run = 1e-3 * rep(draws$run, 2), 
    tauc_run = 1e-3 * c( tauc[tauc$survey_gaps == 1, 'TAUC'],  tauc[tauc$survey_gaps == 7, 'TAUC'])  ) 
ggplot(plt_dat, aes(x = draw_run, y=tauc_run, colour = survey_gaps)) +
  theme_bw() + theme(panel.grid.minor = element_blank()) +
  geom_point(size = 2) +
  geom_abline(slope=1, intercept=0, color='red') +
  labs(x=" True Run ('000)",y="Run from TAUC ('000)") 
ggsave('TAUC run vs draw run.pdf')
```
